---
title: "GSE164378 RNA-ADT Debug: Fixed-Direction Quantities"
output: html_document
---

This is a debug page, not a tutorial. It keeps the setup deliberately raw:

- use the full 4,000-cell RNA fit only to define fixed RNA feature directions;
- sample nested cell sets, so each smaller `n` is contained in larger `n`;
- stratify by RNA molecule sampling rate `rho`;
- for each `(rho, n)`, compute variance along the fixed directions, `d_X^2`,
  `q_X`, `theta_X`, and RNA-to-ADT `I`.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

rna_counts <- gse164378_3p_citeseq_hvg$rna_counts
adt_counts <- gse164378_3p_citeseq_hvg$adt_counts
meta <- gse164378_3p_citeseq_hvg$meta

r_max <- 12
count_transform <- "log1p_cpm"
n_grid <- c(500, 1000, 2000, 3000, 4000)
rho_grid <- c(0.35, 0.50, 0.70, 0.85, 1.00)

dim(rna_counts)
dim(adt_counts)
table(meta$celltype.l1)
```

## Reference Directions

```{r reference}
rna_fit <- scscale_fit(
  rna_counts,
  count_transform = count_transform,
  r = r_max,
  fit_umi = FALSE,
  mp_max_iter = 300,
  mp_grid_n = 2500,
  store_matrix = TRUE
)

adt_fit <- scscale_fit(
  adt_counts,
  count_transform = count_transform,
  r = r_max,
  fit_umi = FALSE,
  mp_max_iter = 300,
  mp_grid_n = 2500
)

r <- max(1L, min(
  r_max,
  sum(rna_fit$spikes$is_spike),
  sum(adt_fit$spikes$is_spike),
  length(rna_fit$theta_X),
  length(adt_fit$theta_X)
))
u_ref <- svd(rna_fit$matrix, nu = r, nv = 0)$u[, seq_len(r), drop = FALSE]
theta_Y <- adt_fit$theta_X[seq_len(r)]

data.frame(
  modality = c("RNA", "ADT"),
  p = c(rna_fit$p, adt_fit$p),
  n = c(rna_fit$n, adt_fit$n),
  c_X = c(rna_fit$c_X, adt_fit$c_X),
  n_spikes = c(sum(rna_fit$spikes$is_spike), sum(adt_fit$spikes$is_spike)),
  r = r,
  tau2 = c(rna_fit$bulk$tau2, adt_fit$bulk$tau2),
  lambda_plus = c(rna_fit$bulk$lambda_plus, adt_fit$bulk$lambda_plus),
  selected_iteration = c(rna_fit$bulk$selected_iteration, adt_fit$bulk$selected_iteration)
)
```

## Raw Grid

```{r compute-grid}
set.seed(164378)
cell_order <- sample(colnames(rna_counts), max(n_grid))

rows <- vector("list", length(rho_grid) * length(n_grid))
row_i <- 0L

for (rho in rho_grid) {
  for (n_i in n_grid) {
    row_i <- row_i + 1L
    cells <- cell_order[seq_len(n_i)]
    rna_sub <- rna_counts[, cells, drop = FALSE]
    if (rho < 1) {
      rna_sub <- scscale_downsample_counts_fraction(
        rna_sub,
        fraction = rho,
        seed = 264378 + row_i
      )
    }

    X_sub <- scscale_normalize_counts(
      rna_sub,
      count_transform = count_transform,
      center = TRUE,
      scale = FALSE
    )
    fit_sub <- scscale_fit(
      X_sub,
      input = "normalized",
      center = FALSE,
      scale = FALSE,
      r = r,
      fit_umi = FALSE,
      mp_max_iter = 300,
      mp_grid_n = 1200
    )

    projected <- crossprod(u_ref, X_sub)
    v_fixed <- rowSums(projected^2) / ncol(X_sub)
    d2_X <- pmax(v_fixed - fit_sub$bulk$tau2, 0) / fit_sub$bulk$tau2
    c_X <- nrow(X_sub) / ncol(X_sub)
    q_X <- c_X * d2_X
    theta_X <- scscale_recoverability(q_X, c_X = c_X)
    I <- scscale_low_rank_mi(theta_X, theta_Y)$mi

    rows[[row_i]] <- data.frame(
      sampling_rate = rho,
      n = n_i,
      p = nrow(X_sub),
      c_X = c_X,
      total_umi_observed = sum(rna_sub),
      tau2 = fit_sub$bulk$tau2,
      lambda_plus = fit_sub$bulk$lambda_plus,
      selected_iteration = fit_sub$bulk$selected_iteration,
      I = I,
      rank = seq_len(r),
      v_fixed = v_fixed,
      d2_X = d2_X,
      q_X = q_X,
      theta_X = theta_X
    )
  }
}

debug_grid <- do.call(rbind, rows)
debug_summary <- unique(debug_grid[, c(
  "sampling_rate", "n", "p", "c_X", "total_umi_observed",
  "tau2", "lambda_plus", "selected_iteration", "I"
)])

debug_summary
```

## Cell Mix

```{r cell-mix, fig.width=8, fig.height=4.2}
cell_mix <- do.call(rbind, lapply(n_grid, function(n_i) {
  cells <- cell_order[seq_len(n_i)]
  tab <- prop.table(table(meta[cells, "celltype.l1"]))
  data.frame(n = n_i, celltype = names(tab), fraction = as.numeric(tab))
}))

cell_mix

mix_mat <- xtabs(fraction ~ n + celltype, data = cell_mix)
barplot(
  t(mix_mat),
  beside = FALSE,
  col = grDevices::hcl.colors(ncol(mix_mat), "Dark 3"),
  xlab = "Nested sampled cell number n",
  ylab = "Cell type fraction",
  main = "Cell mix across nested n"
)
legend("right", legend = colnames(mix_mat), fill = grDevices::hcl.colors(ncol(mix_mat), "Dark 3"), bty = "n", cex = 0.8)
```

## Quantities By n, Stratified By U

```{r plot-raw-quantities, fig.width=11, fig.height=14}
rank_cols <- grDevices::hcl.colors(r, "Dark 3")
plot_quantity <- function(y, ylab, main_prefix) {
  old_par <- par(mfrow = c(length(rho_grid), 1), mar = c(3.5, 4.2, 2.2, 1))
  on.exit(par(old_par), add = TRUE)
  for (rho in rho_grid) {
    df_rho <- debug_grid[debug_grid$sampling_rate == rho, ]
    ylim <- range(df_rho[[y]], finite = TRUE)
    plot(
      NA,
      xlim = range(n_grid),
      ylim = ylim,
      xlab = "Nested sampled cell number n",
      ylab = ylab,
      main = paste0(main_prefix, ", rho=", rho)
    )
    for (k in seq_len(r)) {
      df <- df_rho[df_rho$rank == k, ]
      df <- df[order(df$n), ]
      lines(df$n, df[[y]], col = rank_cols[k], lwd = 1.6)
      points(df$n, df[[y]], col = rank_cols[k], pch = 16, cex = 0.65)
    }
    abline(v = 2000, col = "grey40", lty = 2)
  }
}

plot_quantity("v_fixed", expression(v[X*k]), "Fixed-direction variance")
plot_quantity("d2_X", expression(hat(d)[X*k]^2), "d2")
plot_quantity("q_X", expression(q[X*k]), "q")
plot_quantity("theta_X", expression(theta[X*k]), "theta")
```

## I By n

```{r plot-I, fig.width=8, fig.height=4.8}
I_cols <- grDevices::hcl.colors(length(rho_grid), "Dark 3")
plot(
  NA,
  xlim = range(n_grid),
  ylim = range(debug_summary$I),
  xlab = "Nested sampled cell number n",
  ylab = expression(I),
  main = "RNA-ADT I by n, stratified by rho"
)
for (i in seq_along(rho_grid)) {
  rho <- rho_grid[i]
  df <- debug_summary[debug_summary$sampling_rate == rho, ]
  df <- df[order(df$n), ]
  lines(df$n, df$I, col = I_cols[i], lwd = 2)
  points(df$n, df$I, col = I_cols[i], pch = 16)
}
abline(v = 2000, col = "grey40", lty = 2)
legend(
  "topright",
  legend = paste0("rho=", rho_grid),
  col = I_cols,
  lwd = 2,
  pch = 16,
  bty = "n"
)
```
