---
title: "Spike Model Versus Direct Scaling Fit With GSE164378"
output: html_document
---

This tutorial compares two ways to model RNA-ADT mutual-information scaling:

1. `scscale()` derives \(I(n,U)\) from the fitted Gaussian spikes.
2. `scscale_empirical_scaling_fit()` fits a scaling law directly to sampled
   empirical MI values.

Both methods use the same RNA, ADT, cell-number, and UMI-depth ranges.

```{r setup, include=FALSE}
library(scScale)
data(gse164378_3p_citeseq_hvg)

X <- gse164378_3p_citeseq_hvg$rna_counts
Y <- gse164378_3p_citeseq_hvg$adt_counts
n_grid <- c(500, 1000, 2000, 4000)
```

## Fit the Two Models

```{r fit-models}
spike_model <- scscale(
  X,
  Y,
  n_grid = n_grid,
  seed = 164378
)

direct_model <- scscale_empirical_scaling_fit(
  X,
  Y,
  n_grid = n_grid,
  sampling_rates = spike_model$grids$umi_sampling_rates,
  r = 12,
  cell_replicates = 2,
  n_workers = 2,
  joint_mode = "hybrid",
  seed = 264378
)
```

The spike model uses the MP/BBP spectral fit. The direct model samples cells
and counts, measures empirical subspace MI, and fits smooth marginal and joint
scaling laws to those measurements.

```{r prepare-comparison, results='hide'}
empirical <- scscale_empirical_values(direct_model, summarize = TRUE)
empirical$n <- empirical$cell_sampling_rate * direct_model$sampling$n_total

rates <- sort(spike_model$grids$umi_sampling_rates)
rate_dense <- seq(min(rates), max(rates), length.out = 200)
n_dense <- seq(min(n_grid), max(n_grid), length.out = 200)

direct_umi <- scscale_predict_empirical_umi(
  direct_model,
  sampling_rate = rate_dense
)
direct_cell <- scscale_predict_empirical_cell(
  direct_model,
  n = n_dense
)
direct_joint <- scscale_predict_empirical_joint(
  direct_model,
  n = n_dense,
  sampling_rate = rates
)
```

## Model Comparison

The same empirical measurements are overlaid in both columns. The left column
shows the spike model; the right column shows the direct scaling-law fit.

```{r comparison-plots, fig.width=12, fig.height=12}
draw_error_bars <- function(x, y, sd, col, length = 0.04) {
  keep <- is.finite(x) & is.finite(y) & is.finite(sd) & sd > 1e-8
  if (!any(keep)) return(invisible(NULL))
  arrows(
    x[keep],
    y[keep] - sd[keep],
    x[keep],
    y[keep] + sd[keep],
    angle = 90,
    code = 3,
    length = length,
    col = col
  )
}

add_empirical <- function(x, y, sd, col = "#D55E00", pch = 17) {
  points(x, y, pch = pch, col = col)
  draw_error_bars(x, y, sd, col = col)
}

rate_cols <- setNames(
  grDevices::hcl.colors(length(rates), palette = "Dark 3"),
  as.character(rates)
)

umi_empirical <- empirical[
  empirical$cell_sampling_rate == max(empirical$cell_sampling_rate),
  ,
  drop = FALSE
]
cell_empirical <- empirical[
  empirical$sampling_rate == max(empirical$sampling_rate),
  ,
  drop = FALSE
]

umi_ylim <- range(
  spike_model$umi$fitted_curve$I_theory,
  spike_model$umi$fitted_curve$I_infinity,
  direct_umi$I_fit,
  umi_empirical$I_empirical,
  finite = TRUE
)
cell_ylim <- range(
  spike_model$cell$I_theory,
  spike_model$pair$I_infinity,
  direct_cell$I_fit,
  cell_empirical$I_empirical,
  finite = TRUE
)
joint_ylim <- range(
  spike_model$joint$I_theory,
  direct_joint$I_fit,
  empirical$I_empirical,
  finite = TRUE
)

old_par <- par(mfrow = c(3, 2), mar = c(4.2, 4.3, 2.5, 1))

plot(
  spike_model,
  type = "umi",
  ylim = umi_ylim,
  main = "UMI: spike model"
)
add_empirical(
  umi_empirical$sampling_rate,
  umi_empirical$I_empirical,
  umi_empirical$I_sd
)

plot(
  direct_umi$sampling_rate,
  direct_umi$I_fit,
  type = "l",
  lwd = 2,
  col = "#CC79A7",
  ylim = umi_ylim,
  xlab = expression("UMI sampling rate " * rho),
  ylab = expression(I),
  main = "UMI: direct scaling fit"
)
add_empirical(
  umi_empirical$sampling_rate,
  umi_empirical$I_empirical,
  umi_empirical$I_sd
)
legend(
  "bottomright",
  legend = c("direct fit", "empirical"),
  col = c("#CC79A7", "#D55E00"),
  lwd = c(2, NA),
  pch = c(NA, 17),
  bty = "n"
)

plot(
  spike_model,
  type = "cell",
  ylim = cell_ylim,
  main = "Cell number: spike model"
)
add_empirical(
  cell_empirical$n,
  cell_empirical$I_empirical,
  cell_empirical$I_sd
)

plot(
  direct_cell$n,
  direct_cell$I_fit,
  type = "l",
  lwd = 2,
  col = "#CC79A7",
  ylim = cell_ylim,
  xlab = "Cell number n",
  ylab = expression(I),
  main = "Cell number: direct scaling fit"
)
add_empirical(
  cell_empirical$n,
  cell_empirical$I_empirical,
  cell_empirical$I_sd
)
legend(
  "bottomright",
  legend = c("direct fit", "empirical"),
  col = c("#CC79A7", "#D55E00"),
  lwd = c(2, NA),
  pch = c(NA, 17),
  bty = "n"
)

plot(
  spike_model,
  type = "joint",
  ylim = joint_ylim,
  main = expression("Joint " * I(n, U) * ": spike model")
)
for (rho in rates) {
  observed <- empirical[empirical$sampling_rate == rho, , drop = FALSE]
  col <- rate_cols[[as.character(rho)]]
  points(observed$n, observed$I_empirical, pch = 1, col = col)
  draw_error_bars(
    observed$n,
    observed$I_empirical,
    observed$I_sd,
    col = col,
    length = 0.03
  )
}

plot(
  NA_real_,
  xlim = range(n_grid),
  ylim = joint_ylim,
  xlab = "Cell number n",
  ylab = expression(I(n, rho)),
  main = expression("Joint " * I(n, U) * ": direct scaling fit")
)
for (rho in rates) {
  fitted <- direct_joint[
    direct_joint$sampling_rate == rho,
    ,
    drop = FALSE
  ]
  observed <- empirical[
    empirical$sampling_rate == rho,
    ,
    drop = FALSE
  ]
  col <- rate_cols[[as.character(rho)]]
  lines(fitted$n, fitted$I_fit, col = col, lwd = 2)
  points(observed$n, observed$I_empirical, pch = 1, col = col)
  draw_error_bars(
    observed$n,
    observed$I_empirical,
    observed$I_sd,
    col = col,
    length = 0.03
  )
}
legend(
  "bottomright",
  legend = paste0("rho=", rates),
  col = rate_cols,
  lwd = 2,
  bty = "n"
)

par(old_par)
```
