---
title: "Direct Empirical Scaling Fit With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial runs the direct empirical scaling workflow on the same
GSE164378 3' CITE-seq RNA-ADT example used by the Gaussian spike tutorials.
Here we do not fit the spike model. We sample a grid of cell numbers and UMI
depths, compute fixed-rank empirical RNA-ADT mutual information, and fit the
empirical scaling law directly.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

rna_counts <- gse164378_3p_citeseq_hvg$rna_counts
adt_counts <- gse164378_3p_citeseq_hvg$adt_counts
count_transform <- "log1p_cpm"

joint_plot_rates <- c(0.20, 0.50, 1.00)
r_empirical <- 12
available_cores <- parallel::detectCores(logical = TRUE)
n_workers <- if (is.na(available_cores)) 2L else max(1L, min(4L, available_cores))
```

## Fit Empirical Scaling Law

`scscale_empirical_scaling_fit()` samples the grid, preprocesses each sampled
matrix in the standard way, computes empirical subspace MI, and fits the
cell-number, UMI, and joint laws into one object.

```{r fit-empirical-scaling}
fit <- scscale_empirical_scaling_fit(
  rna_counts,
  adt_counts,
  r = r_empirical,
  joint_mode = "hybrid",
  seed = 164378,
  count_transform = count_transform,
  n_workers = n_workers
)

fit
```

The fitted joint law uses one global parameter set across the sampled grid.

```{r prepare-plot-data}
grid <- fit$grid
sampling_rates <- sort(unique(grid$sampling_rate))
cell_sampling_rates <- sort(unique(grid$cell_sampling_rate))
grid$sampling_rate_label <- factor(
  grid$sampling_rate,
  levels = sampling_rates,
  labels = paste0("rho = ", sampling_rates)
)

rho_to_U <- aggregate(U ~ sampling_rate, data = grid, FUN = unique)
rho_to_U <- rho_to_U[order(rho_to_U$sampling_rate), ]
rho_cols <- setNames(
  c("#0072B2", "#56B4E9", "#D55E00", "#E69F00", "#009E73", "#CC79A7", "#999999", "#8B5A2B", "#000000"),
  as.character(rho_to_U$sampling_rate)
)

summarize_mi <- function(data, by) {
  mean_df <- aggregate(
    stats::as.formula(paste("I_empirical ~", paste(by, collapse = " + "))),
    data = data,
    FUN = mean
  )
  sd_df <- aggregate(
    stats::as.formula(paste("I_empirical ~", paste(by, collapse = " + "))),
    data = data,
    FUN = stats::sd
  )
  n_df <- aggregate(
    stats::as.formula(paste("replicate ~", paste(by, collapse = " + "))),
    data = data,
    FUN = function(x) length(unique(x))
  )
  mean_df$I_sd <- sd_df$I_empirical
  mean_df$n_replicates <- n_df$replicate
  mean_df
}

draw_vertical_bars <- function(x, y, sd, col, cap = NULL, lwd = 1.6) {
  sd[!is.finite(sd)] <- 0
  if (is.null(cap)) {
    cap <- diff(range(x)) * 0.012
  }
  segments(x, y - sd, x, y + sd, col = col, lwd = lwd)
  segments(x - cap, y - sd, x + cap, y - sd, col = col, lwd = lwd)
  segments(x - cap, y + sd, x + cap, y + sd, col = col, lwd = lwd)
}
```

## Cell-Number Scaling

The cell-number marginal fit uses the full-depth grid slice.

```{r cell-number-scaling, fig.width=6.8, fig.height=4.6}
cell_obs <- subset(grid, abs(U - fit$reference$U_ref) < 1e-8)
cell_obs <- summarize_mi(cell_obs, by = "cell_sampling_rate")
cell_obs <- cell_obs[order(cell_obs$cell_sampling_rate), ]

cell_rate_curve <- seq(min(cell_sampling_rates), max(cell_sampling_rates), length.out = 200)
n_curve <- cell_rate_curve * max(grid$n)
cell_curve <- data.frame(cell_sampling_rate = cell_rate_curve, n = n_curve)
cell_curve$I_fit <- predict(fit$cell_fit, cell_curve)

plot(
  NA,
  xlim = range(cell_obs$cell_sampling_rate),
  ylim = range(c(cell_obs$I_empirical - cell_obs$I_sd, cell_obs$I_empirical + cell_obs$I_sd, cell_curve$I_fit), na.rm = TRUE),
  xlab = expression("Cell sampling rate " * gamma),
  ylab = expression(hat(I)(n~"|"~U[ref])),
  main = "Empirical cell-number scaling"
)
draw_vertical_bars(cell_obs$cell_sampling_rate, cell_obs$I_empirical, cell_obs$I_sd, col = "#0072B2")
lines(cell_curve$cell_sampling_rate, cell_curve$I_fit, col = "#D55E00", lwd = 2)
legend(
  "bottomright",
  legend = c("empirical mean +/- sd", "empirical fit"),
  col = c("#0072B2", "#D55E00"),
  pch = c(NA, NA),
  lwd = c(1.6, 2),
  bty = "n"
)
```

## UMI Scaling

The UMI marginal fit uses the full-cell grid slice.

```{r umi-scaling, fig.width=6.8, fig.height=4.6}
umi_obs <- subset(grid, n == fit$reference$n_ref)
umi_obs <- summarize_mi(umi_obs, by = c("sampling_rate", "U"))
umi_obs <- umi_obs[order(umi_obs$sampling_rate), ]

rho_curve <- seq(min(sampling_rates), max(sampling_rates), length.out = 200)
U_curve <- rho_curve * stats::median(Matrix::colSums(rna_counts))
umi_curve <- data.frame(U = U_curve, sampling_rate = rho_curve)
umi_curve$I_fit <- predict(fit$umi_fit, umi_curve)

plot(
  NA,
  xlim = range(umi_obs$sampling_rate),
  ylim = range(c(umi_obs$I_empirical - umi_obs$I_sd, umi_obs$I_empirical + umi_obs$I_sd, umi_curve$I_fit), na.rm = TRUE),
  xlab = expression("Sampling rate " * rho),
  ylab = expression(hat(I)(U~"|"~n[ref])),
  main = "Empirical UMI scaling"
)
draw_vertical_bars(umi_obs$sampling_rate, umi_obs$I_empirical, umi_obs$I_sd, col = "#0072B2")
lines(umi_curve$sampling_rate, umi_curve$I_fit, col = "#D55E00", lwd = 2)
legend(
  "bottomright",
  legend = c("empirical mean +/- sd", "empirical fit"),
  col = c("#0072B2", "#D55E00"),
  pch = c(NA, NA),
  lwd = c(1.6, 2),
  bty = "n"
)
```

## Joint Scaling By UMI Level

The joint fit uses all grid points and one global parameter set. Each line below
is the fitted cell-number scaling curve at one UMI sampling rate.

```{r joint-scaling-by-umi, fig.width=8, fig.height=5.2}
joint_obs <- summarize_mi(grid, by = c("cell_sampling_rate", "sampling_rate", "U"))
plot(
  NA,
  xlim = range(cell_sampling_rates),
  ylim = range(c(joint_obs$I_empirical - joint_obs$I_sd, joint_obs$I_empirical + joint_obs$I_sd, fit$fitted), na.rm = TRUE),
  xlab = expression("Cell sampling rate " * gamma),
  ylab = expression(hat(I)(n,U)),
  main = "Empirical joint scaling by UMI level"
)

for (rho in joint_plot_rates) {
  U_i <- rho_to_U$U[rho_to_U$sampling_rate == rho]
  obs_i <- joint_obs[joint_obs$sampling_rate == rho, ]
  obs_i <- obs_i[order(obs_i$cell_sampling_rate), ]
  draw_vertical_bars(obs_i$cell_sampling_rate, obs_i$I_empirical, obs_i$I_sd, col = rho_cols[as.character(rho)])
  pred_rate <- seq(min(cell_sampling_rates), max(cell_sampling_rates), length.out = 300)
  pred_i <- data.frame(cell_sampling_rate = pred_rate, n = pred_rate * max(grid$n), U = U_i)
  pred_i$I_fit <- predict(fit, pred_i)
  lines(pred_i$cell_sampling_rate, pred_i$I_fit, col = rho_cols[as.character(rho)], lwd = 2)
}

legend(
  "bottomright",
  legend = paste0("rho = ", joint_plot_rates),
  col = rho_cols[as.character(joint_plot_rates)],
  pch = NA,
  lwd = 2,
  bty = "n",
  cex = 0.85
)
```

## Marginal And Joint Fit Check

The two marginal fits should be compatible with the corresponding slices of
the joint fit. Here we compare the cell-number marginal law with the joint law
at full depth, and the UMI marginal law with the joint law at the full cell
number.

```{r marginal-joint-check, fig.width=11, fig.height=4.5}
cell_joint_curve <- data.frame(cell_sampling_rate = cell_rate_curve, n = n_curve, U = fit$reference$U_ref)
cell_joint_curve$I_fit <- predict(fit, cell_joint_curve)

umi_joint_curve <- data.frame(U = U_curve, sampling_rate = rho_curve, n = fit$reference$n_ref)
umi_joint_curve$I_fit <- predict(fit, umi_joint_curve)

old_par <- par(mfrow = c(1, 2), mar = c(4.2, 4.3, 2.8, 1))
plot(
  NA,
  xlim = range(cell_obs$cell_sampling_rate),
  ylim = range(c(cell_obs$I_empirical - cell_obs$I_sd, cell_obs$I_empirical + cell_obs$I_sd, cell_curve$I_fit, cell_joint_curve$I_fit), na.rm = TRUE),
  xlab = expression("Cell sampling rate " * gamma),
  ylab = expression(hat(I)(n~"|"~U[ref])),
  main = "Cell marginal vs joint slice"
)
draw_vertical_bars(cell_obs$cell_sampling_rate, cell_obs$I_empirical, cell_obs$I_sd, col = "#0072B2")
lines(cell_curve$cell_sampling_rate, cell_curve$I_fit, col = "#D55E00", lwd = 2)
lines(cell_joint_curve$cell_sampling_rate, cell_joint_curve$I_fit, col = "#009E73", lwd = 2, lty = 2)
legend(
  "bottomright",
  legend = c("empirical mean +/- sd", "marginal fit", "joint slice"),
  col = c("#0072B2", "#D55E00", "#009E73"),
  pch = c(NA, NA, NA),
  lwd = c(1.6, 2, 2),
  lty = c(NA, 1, 2),
  bty = "n",
  cex = 0.9
)

plot(
  NA,
  xlim = range(umi_obs$sampling_rate),
  ylim = range(c(umi_obs$I_empirical - umi_obs$I_sd, umi_obs$I_empirical + umi_obs$I_sd, umi_curve$I_fit, umi_joint_curve$I_fit), na.rm = TRUE),
  xlab = expression("Sampling rate " * rho),
  ylab = expression(hat(I)(U~"|"~n[ref])),
  main = "UMI marginal vs joint slice"
)
draw_vertical_bars(umi_obs$sampling_rate, umi_obs$I_empirical, umi_obs$I_sd, col = "#0072B2")
lines(umi_curve$sampling_rate, umi_curve$I_fit, col = "#D55E00", lwd = 2)
lines(umi_joint_curve$sampling_rate, umi_joint_curve$I_fit, col = "#009E73", lwd = 2, lty = 2)
legend(
  "bottomright",
  legend = c("empirical mean +/- sd", "marginal fit", "joint slice"),
  col = c("#0072B2", "#D55E00", "#009E73"),
  pch = c(NA, NA, NA),
  lwd = c(1.6, 2, 2),
  lty = c(NA, 1, 2),
  bty = "n",
  cex = 0.9
)
par(old_par)
```
