---
title: "Spike Model Fit Versus Empirical Model Fit With GSE164378"
output: html_document
---

This tutorial compares two ways to model RNA-ADT mutual-information scaling:

1. Fit a spike model with `scscale_spike_fit()`.
2. Fit empirical scaling curves with `scscale_empirical_scaling_fit()`.

The two fits below use the same RNA, ADT, cell-number, and UMI-depth ranges.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(scScale)
data(gse164378_3p_citeseq_hvg)

X <- gse164378_3p_citeseq_hvg$rna_counts
Y <- gse164378_3p_citeseq_hvg$adt_counts
n_grid <- c(500, 1000, 2000, 4000)
rates <- c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)
```

## Fit the Two Models

```{r fit-models, results='hide'}
spike_fit <- scscale_spike_fit(
  X,
  Y,
  umi_sampling_rates = rates,
  x_rank = 12,
  y_rank = 12,
  seed = 164378
)

empirical_fit <- scscale_empirical_scaling_fit(
  X,
  Y,
  n_grid = n_grid,
  sampling_rates = rates,
  r = 12,
  cell_replicates = 2,
  n_workers = 2,
  joint_mode = "hybrid",
  seed = 264378
)
```

```{r prepare-comparison, results='hide'}
empirical <- scscale_empirical_values(empirical_fit, summarize = TRUE)
empirical$n <- empirical$cell_sampling_rate * empirical_fit$sampling$n_total

rate_dense <- seq(min(rates), max(rates), length.out = 200)
n_dense <- seq(min(n_grid), max(n_grid), length.out = 200)

spike_umi <- scscale_umi_scaling(
  spike_fit,
  sampling_rate = rate_dense
)
spike_cell <- scscale_cell_scaling(
  spike_fit,
  n = n_dense
)
spike_joint <- scscale_joint_scaling(
  spike_fit,
  n = n_dense,
  sampling_rate = rates
)

empirical_umi <- scscale_predict_empirical_umi(
  empirical_fit,
  sampling_rate = rate_dense
)
empirical_cell <- scscale_predict_empirical_cell(
  empirical_fit,
  n = n_dense
)
empirical_joint <- scscale_predict_empirical_joint(
  empirical_fit,
  n = n_dense,
  sampling_rate = rates
)
empirical_joint_parameters <- c(
  empirical_fit$joint_fit$coefficients,
  empirical_fit$joint_fit$fixed
)
empirical_I_infinity <- unname(
  empirical_joint_parameters[["I_infinity"]]
)
```

## Model Comparison

The spike-model fits are shown on the left and the empirical curve fits on the
right. Dashed lines show the corresponding shared \(I_\infty\) estimates.

```{r comparison-plots, fig.width=12, fig.height=12}
plot_curve <- function(fit, x, y, ceiling, title, xlab, ylim, observed = NULL) {
  plot(fit[[x]], fit[[y]], type = "l", lwd = 2, col = "#0072B2",
       xlab = xlab, ylab = "MI", main = title, ylim = ylim)
  abline(h = ceiling, lty = 2, col = "grey35", lwd = 2)
  if (!is.null(observed)) {
    keep <- is.finite(observed$I_sd) & observed$I_sd > 0
    segments(observed[[x]][keep], observed$I_empirical[keep] - observed$I_sd[keep],
             observed[[x]][keep], observed$I_empirical[keep] + observed$I_sd[keep],
             col = "#D55E00")
    points(observed[[x]], observed$I_empirical, pch = 16, col = "#D55E00")
  }
}

plot_joint <- function(fit, y, title, ylim, observed = NULL) {
  cols <- setNames(hcl.colors(length(rates), "Dark 3"), rates)
  plot(NA, xlim = range(n_grid), ylim = ylim, xlab = "Cell number n",
       ylab = "MI", main = title)
  for (rho in rates) {
    z <- fit[fit$sampling_rate == rho, ]
    lines(z$n, z[[y]], col = cols[as.character(rho)], lwd = 2)
  }
  if (!is.null(observed)) {
    point_cols <- cols[as.character(observed$sampling_rate)]
    keep <- is.finite(observed$I_sd) & observed$I_sd > 0
    segments(observed$n[keep], observed$I_empirical[keep] - observed$I_sd[keep],
             observed$n[keep], observed$I_empirical[keep] + observed$I_sd[keep],
             col = point_cols[keep])
    points(observed$n, observed$I_empirical, pch = 1,
           col = point_cols)
  }
  legend("bottomright", paste0("rho=", rates), col = cols, lwd = 2, bty = "n")
}

umi_observed <- subset(empirical, cell_sampling_rate == max(cell_sampling_rate))
cell_observed <- subset(empirical, sampling_rate == max(sampling_rate))
umi_ylim <- range(spike_umi$I_theory, spike_fit$I_infinity,
                  empirical_umi$I_fit, empirical_I_infinity, finite = TRUE)
cell_ylim <- range(spike_cell$I_theory, spike_fit$I_infinity,
                   empirical_cell$I_fit, empirical_I_infinity, finite = TRUE)
joint_ylim <- range(spike_joint$I_theory, empirical_joint$I_fit, finite = TRUE)

old_par <- par(mfrow = c(3, 2), mar = c(4.2, 4.3, 2.5, 1))
plot_curve(spike_umi, "sampling_rate", "I_theory", spike_fit$I_infinity,
           "UMI: spike model", "UMI sampling rate", umi_ylim)
plot_curve(empirical_umi, "sampling_rate", "I_fit", empirical_I_infinity,
           "UMI: empirical model fit", "UMI sampling rate", umi_ylim, umi_observed)
plot_curve(spike_cell, "n", "I_theory", spike_fit$I_infinity,
           "Cell number: spike model", "Cell number n", cell_ylim)
plot_curve(empirical_cell, "n", "I_fit", empirical_I_infinity,
           "Cell number: empirical model fit", "Cell number n", cell_ylim, cell_observed)
plot_joint(spike_joint, "I_theory", "Joint: spike model", joint_ylim)
plot_joint(empirical_joint, "I_fit", "Joint: empirical model fit",
           joint_ylim, empirical)
par(old_par)
```
