---
title: "RNA-ADT Mutual Information With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial uses the paired GSE164378 3' CITE-seq example to estimate
RNA-to-ADT spectral mutual information. RNA is the measured modality `X`; ADT
is the paired target modality `Y`.

```{r setup, include=FALSE}
library(scScale)
data(gse164378_3p_citeseq_hvg)

rna_counts <- gse164378_3p_citeseq_hvg$rna_counts
adt_counts <- gse164378_3p_citeseq_hvg$adt_counts
```

## Fit RNA and ADT

```{r fit-modalities}
r <- 12

rna_fit <- scscale_fit(
  rna_counts,
  r = r,
  fit_umi = FALSE,
  mp_max_iter = 20,
  mp_grid_n = 4000,
  store_matrix = TRUE
)

adt_fit <- scscale_fit(
  adt_counts,
  r = r,
  fit_umi = FALSE,
  mp_max_iter = 20,
  mp_grid_n = 4000
)

data.frame(
  modality = c("RNA", "ADT"),
  p = c(rna_fit$p, adt_fit$p),
  n = c(rna_fit$n, adt_fit$n),
  c_X = c(rna_fit$c_X, adt_fit$c_X),
  tau2 = c(rna_fit$bulk$tau2, adt_fit$bulk$tau2),
  lambda_plus = c(rna_fit$bulk$lambda_plus, adt_fit$bulk$lambda_plus),
  selected_iteration = c(rna_fit$bulk$selected_iteration, adt_fit$bulk$selected_iteration)
)

theta_Y <- adt_fit$theta_X[seq_len(r)]

data.frame(
  I_theory = scscale_theory_mi(rna_fit, adt_fit, r = r)$I_theory,
  I_infinity = scscale_i_infinity(rna_fit, adt_fit, r = r)$I_infinity
)
```

## UMI Scaling Law

```{r fit-umi-mi}
sampling_rates <- c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)

umi_scaling <- scscale_umi_scaling(
  rna_counts,
  sampling_rates = sampling_rates,
  r = r,
  seed = 164378,
  n_replicates = 1,
  theta_Y = theta_Y,
  reference_fit = rna_fit,
  empirical = FALSE,
  mp_max_iter = 20,
  mp_grid_n = 2500
)

umi_mi <- unique(umi_scaling$scaling[, c(
  "sampling_rate", "U", "total_umi_observed", "I_theory", "I_infinity"
)])

umi_mi
umi_scaling$q_fit_rate
```

## Cell Scaling Curves

For a fixed sequencing rate `rho`, the fitted RNA spike strengths give
`d2_X(rho) = q_X(rho) / c_ref`, where `c_ref = p / n_ref` at 4,000 cells.
Changing cell number changes `c_X = p / n`, while the fixed-UMI spike strengths
are held on that sequencing-rate curve.

```{r fit-cell-mi}
n_grid <- c(500, 1000, 2000, 3000, 4000)
rho_grid <- c(0.35, 0.50, 0.70, 0.85, 1.00)
c_ref <- rna_fit$c_X

cell_theta <- do.call(rbind, lapply(rho_grid, function(rho) {
  q_df <- unique(umi_scaling$scaling[umi_scaling$scaling$sampling_rate == rho,
                                    c("rank", "q_X_rate_hat")])
  q_df <- q_df[order(q_df$rank), ]
  d2_rho <- q_df$q_X_rate_hat / c_ref
  out <- scscale_cell_scaling(d2_rho, p = rna_fit$p, n_grid = n_grid, theta_Y = theta_Y)
  transform(out, sampling_rate = rho)
}))

cell_curves <- aggregate(
  I_theory ~ n + c_X + sampling_rate,
  data = cell_theta,
  FUN = unique
)

cell_curves
```

## Recoverability Check

```{r plot-theta-cell-scaling, fig.width=7, fig.height=4.2}
theta_full <- cell_theta[cell_theta$sampling_rate == 1 & cell_theta$rank <= 6, ]
theta_cols <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#56B4E9", "#E69F00")

plot(
  NA,
  xlim = range(theta_full$n),
  ylim = range(theta_full$theta_X),
  xlab = "Cell number n",
  ylab = expression(theta[X*k](n)),
  main = expression("Recoverability at " * rho == 1)
)
for (k in sort(unique(theta_full$rank))) {
  df <- theta_full[theta_full$rank == k, ]
  df <- df[order(df$n), ]
  lines(df$n, df$theta_X, col = theta_cols[k], lwd = 2)
  points(df$n, df$theta_X, col = theta_cols[k], pch = 16)
}
legend(
  "bottomleft",
  legend = paste0("k=", 1:6),
  col = theta_cols,
  lwd = 2,
  pch = 16,
  bty = "n",
  cex = 0.85
)
```

## Empirical d2 Check

If the population spike strength `d_X^2` is stable, the estimates below should
be roughly flat across sampled cell numbers at a fixed sequencing rate.

```{r empirical-d2-cell-scaling}
d2_grid <- expand.grid(
  sampling_rate = rho_grid,
  n = n_grid,
  KEEP.OUT.ATTRS = FALSE
)

u_ref <- svd(rna_fit$matrix, nu = r, nv = 0)$u[, seq_len(r), drop = FALSE]
d2_rows <- vector("list", nrow(d2_grid))
for (i in seq_len(nrow(d2_grid))) {
  rho <- d2_grid$sampling_rate[i]
  n_i <- d2_grid$n[i]
  set.seed(364378 + i)
  cells <- sample(colnames(rna_counts), n_i)

  rna_sub <- rna_counts[, cells, drop = FALSE]
  if (rho < 1) {
    rna_sub <- scscale_downsample_counts_fraction(rna_sub, fraction = rho, seed = 464378 + i)
  }

  X_sub <- scscale_normalize_counts(rna_sub, center = TRUE, scale = TRUE)
  fit_sub <- scscale_fit(
    X_sub,
    input = "normalized",
    center = FALSE,
    scale = FALSE,
    r = r,
    mp_max_iter = 20,
    mp_grid_n = 1500
  )

  projected <- crossprod(u_ref, X_sub)
  v <- rowSums(projected^2) / ncol(X_sub)
  d2_hat <- pmax(v - fit_sub$bulk$tau2, 0) / fit_sub$bulk$tau2
  d2_rows[[i]] <- data.frame(
    sampling_rate = rho,
    n = n_i,
    rank = seq_len(r),
    tau2 = fit_sub$bulk$tau2,
    d2_X = d2_hat
  )
}

d2_empirical <- do.call(rbind, d2_rows)
d2_empirical[d2_empirical$rank <= 6, ]
```

```{r plot-d2-cell-scaling, fig.width=11, fig.height=3.7}
d2_ranks <- 1:6
d2_plot <- d2_empirical[d2_empirical$rank %in% d2_ranks, ]

old_par <- par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2.4, 1))
for (rho in c(0.35, 0.70, 1.00)) {
  df_rho <- d2_plot[d2_plot$sampling_rate == rho, ]
  plot(
    NA,
    xlim = range(df_rho$n),
    ylim = range(df_rho$d2_X),
    xlab = "Sampled cell number n",
    ylab = expression(hat(d)[X*k]^2),
    main = bquote(rho == .(rho))
  )
  for (k in d2_ranks) {
    df <- df_rho[df_rho$rank == k, ]
    df <- df[order(df$n), ]
    lines(df$n, df$d2_X, col = theta_cols[k], lwd = 2)
    points(df$n, df$d2_X, col = theta_cols[k], pch = 16)
  }
}
par(old_par)
```

## Empirical Comparison

The empirical points recompute RNA and ADT cell subspaces on matched cell
subsamples. RNA is also globally thinned at the specified sequencing rate.

```{r empirical-mi}
empirical_grid <- expand.grid(
  sampling_rate = rho_grid,
  n = n_grid,
  KEEP.OUT.ATTRS = FALSE
)

empirical_rows <- vector("list", nrow(empirical_grid))
for (i in seq_len(nrow(empirical_grid))) {
  rho <- empirical_grid$sampling_rate[i]
  n_i <- empirical_grid$n[i]
  set.seed(164378 + i)
  cells <- sample(colnames(rna_counts), n_i)

  rna_sub <- rna_counts[, cells, drop = FALSE]
  adt_sub <- adt_counts[, cells, drop = FALSE]
  if (rho < 1) {
    rna_sub <- scscale_downsample_counts_fraction(rna_sub, fraction = rho, seed = 264378 + i)
  }

  X <- scscale_normalize_counts(rna_sub, center = TRUE, scale = TRUE)
  Y <- scscale_normalize_counts(adt_sub, center = TRUE, scale = TRUE)
  z_X <- right_singular_vectors(X, r = r)
  z_Y <- right_singular_vectors(Y, r = r)
  mi_emp <- subspace_overlap_mi(z_X, z_Y)

  theory_match <- cell_curves[
    cell_curves$sampling_rate == rho & cell_curves$n == n_i,
    "I_theory"
  ]
  empirical_rows[[i]] <- data.frame(
    sampling_rate = rho,
    n = n_i,
    I_empirical = mi_emp$mi,
    I_theory = theory_match,
    r_eff = mi_emp$r_eff
  )
}

empirical_mi <- do.call(rbind, empirical_rows)
empirical_mi
```

## Plots

```{r plot-mi, fig.width=11, fig.height=7.2}
cols <- c("0.35" = "#0072B2", "0.50" = "#D55E00", "0.70" = "#009E73",
          "0.85" = "#CC79A7", "1" = "#E69F00")

old_par <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.4, 1))

plot(
  umi_mi$sampling_rate,
  umi_mi$I_theory,
  pch = 16,
  col = "#0072B2",
  type = "b",
  xlab = expression("Sampling rate " * rho),
  ylab = expression(I(rho)),
  main = "UMI scaling law"
)
lines(umi_mi$sampling_rate, umi_mi$I_infinity, col = "#D55E00", lwd = 2, lty = 2)
legend(
  "bottomright",
  legend = c("finite n", "infinite n"),
  col = c("#0072B2", "#D55E00"),
  pch = c(16, NA),
  lwd = c(1, 2),
  lty = c(1, 2),
  bty = "n"
)

plot(
  NA,
  xlim = range(n_grid),
  ylim = range(c(cell_curves$I_theory, empirical_mi$I_empirical)),
  xlab = "Cell number n",
  ylab = expression(I(n, rho)),
  main = "I(n, U) at fixed UMI rates"
)
for (rho in rho_grid) {
  df <- cell_curves[cell_curves$sampling_rate == rho, ]
  df <- df[order(df$n), ]
  lines(df$n, df$I_theory, col = cols[as.character(rho)], lwd = 2)
  emp <- empirical_mi[empirical_mi$sampling_rate == rho, ]
  points(emp$n, emp$I_empirical, pch = 16, col = cols[as.character(rho)])
}
legend(
  "bottomright",
  legend = paste0("rho=", rho_grid),
  col = cols[as.character(rho_grid)],
  lwd = 2,
  pch = 16,
  bty = "n",
  cex = 0.85
)

plot(
  n_grid,
  cell_curves[cell_curves$sampling_rate == 1, "I_theory"],
  type = "b",
  pch = 16,
  col = "#0072B2",
  xlab = "Cell number n",
  ylab = expression(I(n, rho == 1)),
  main = "Cell number scaling law"
)
emp_full <- empirical_mi[empirical_mi$sampling_rate == 1, ]
points(emp_full$n, emp_full$I_empirical, pch = 17, col = "#D55E00")
legend(
  "bottomright",
  legend = c("theory", "empirical"),
  col = c("#0072B2", "#D55E00"),
  pch = c(16, 17),
  lwd = c(1, NA),
  bty = "n"
)

plot(
  empirical_mi$I_theory,
  empirical_mi$I_empirical,
  pch = 16,
  col = cols[as.character(empirical_mi$sampling_rate)],
  xlab = "Theory",
  ylab = "Empirical",
  main = "Empirical comparison"
)
abline(0, 1, col = "black", lwd = 2, lty = 2)
par(old_par)
```
