---
title: "Direct Empirical Scaling Fit With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial runs the direct empirical scaling workflow on a compact subset of
the GSE164378 3' CITE-seq PBMC example. Unlike the Gaussian spike workflow, this
fit does not call spikes or fit an MP bulk. It samples an \(I(n,U)\) grid,
computes fixed-rank empirical RNA-ADT mutual information at each grid point, and
fits one global scaling surface.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

rna_counts <- gse164378_3p_citeseq_hvg$rna_counts[1:800, 1:1500]
adt_counts <- gse164378_3p_citeseq_hvg$adt_counts[, colnames(rna_counts)]

n_grid <- c(300, 600, 900)
U_grid <- c(1000, 2000, 3000)
r_empirical <- 6
```

## Fit Empirical Scaling Law

`scscale_empirical_scaling_fit()` is the all-in-one direct empirical workflow.
It samples cells and UMI depth, preprocesses each sampled matrix in the standard
way, computes empirical subspace MI, and fits the cell-number, UMI, and joint
laws.

```{r fit-empirical-scaling}
fit <- scscale_empirical_scaling_fit(
  rna_counts,
  adt_counts,
  n_grid = n_grid,
  U_grid = U_grid,
  r = r_empirical,
  n_replicates = 2,
  n_ref = max(n_grid),
  U_ref = max(U_grid),
  joint_mode = "hybrid",
  seed = 164378,
  count_transform = "log1p_cpm"
)

fit
```

The fitted joint law uses one global parameter set across the grid.

```{r fitted-parameters}
data.frame(
  parameter = names(c(fit$joint_fit$coefficients, fit$joint_fit$fixed)),
  value = as.numeric(c(fit$joint_fit$coefficients, fit$joint_fit$fixed)),
  row.names = NULL
)

fit$ceiling
```

## Cell-Number Scaling

The cell-number marginal fit uses the grid slice at the reference UMI depth.

```{r cell-number-scaling, fig.width=6.5, fig.height=4.5}
grid <- fit$grid
cell_obs <- aggregate(I_empirical ~ n, data = subset(grid, U == fit$reference$U_ref), FUN = mean)
cell_sd <- aggregate(I_empirical ~ n, data = subset(grid, U == fit$reference$U_ref), FUN = stats::sd)
cell_obs$I_sd <- cell_sd$I_empirical

n_curve <- seq(min(n_grid), max(n_grid), length.out = 100)
cell_curve <- data.frame(n = n_curve)
cell_curve$I_fit <- predict(fit$cell_fit, cell_curve)

plot(
  cell_obs$n,
  cell_obs$I_empirical,
  pch = 16,
  col = "#0072B2",
  xlab = "Cell number n",
  ylab = expression(hat(I)(n~"|"~U[ref])),
  main = "Cell-number scaling"
)
arrows(
  cell_obs$n,
  cell_obs$I_empirical - cell_obs$I_sd,
  cell_obs$n,
  cell_obs$I_empirical + cell_obs$I_sd,
  angle = 90,
  code = 3,
  length = 0.04,
  col = "#0072B2"
)
lines(cell_curve$n, cell_curve$I_fit, col = "#D55E00", lwd = 2)
legend(
  "bottomright",
  legend = c("empirical grid mean", "marginal fit"),
  col = c("#0072B2", "#D55E00"),
  pch = c(16, NA),
  lwd = c(NA, 2),
  bty = "n"
)
```

## UMI Scaling

The UMI marginal fit uses the grid slice at the reference cell number.

```{r umi-scaling, fig.width=6.5, fig.height=4.5}
umi_obs <- aggregate(I_empirical ~ U, data = subset(grid, n == fit$reference$n_ref), FUN = mean)
umi_sd <- aggregate(I_empirical ~ U, data = subset(grid, n == fit$reference$n_ref), FUN = stats::sd)
umi_obs$I_sd <- umi_sd$I_empirical

U_curve <- seq(min(U_grid), max(U_grid), length.out = 100)
umi_curve <- data.frame(U = U_curve)
umi_curve$I_fit <- predict(fit$umi_fit, umi_curve)

plot(
  umi_obs$U,
  umi_obs$I_empirical,
  pch = 16,
  col = "#0072B2",
  xlab = "Median UMI per cell U",
  ylab = expression(hat(I)(U~"|"~n[ref])),
  main = "UMI scaling"
)
arrows(
  umi_obs$U,
  umi_obs$I_empirical - umi_obs$I_sd,
  umi_obs$U,
  umi_obs$I_empirical + umi_obs$I_sd,
  angle = 90,
  code = 3,
  length = 0.04,
  col = "#0072B2"
)
lines(umi_curve$U, umi_curve$I_fit, col = "#D55E00", lwd = 2)
legend(
  "bottomright",
  legend = c("empirical grid mean", "marginal fit"),
  col = c("#0072B2", "#D55E00"),
  pch = c(16, NA),
  lwd = c(NA, 2),
  bty = "n"
)
```

## Joint Scaling

The joint fit uses all grid points and one global parameter set. The plot below
shows a cell-number scaling curve at each UMI level.

```{r joint-scaling-by-umi, fig.width=7.5, fig.height=5}
joint_obs <- aggregate(I_empirical ~ n + U, data = grid, FUN = mean)
joint_sd <- aggregate(I_empirical ~ n + U, data = grid, FUN = stats::sd)
joint_obs$I_sd <- joint_sd$I_empirical

cols <- setNames(c("#0072B2", "#D55E00", "#009E73"), as.character(U_grid))
plot(
  NA,
  xlim = range(n_grid),
  ylim = range(c(joint_obs$I_empirical, fit$fitted), na.rm = TRUE),
  xlab = "Cell number n",
  ylab = expression(hat(I)(n,U)),
  main = "Joint empirical scaling by UMI level"
)

for (U_i in U_grid) {
  obs_i <- joint_obs[joint_obs$U == U_i, ]
  obs_i <- obs_i[order(obs_i$n), ]
  points(obs_i$n, obs_i$I_empirical, pch = 16, col = cols[as.character(U_i)])
  arrows(
    obs_i$n,
    obs_i$I_empirical - obs_i$I_sd,
    obs_i$n,
    obs_i$I_empirical + obs_i$I_sd,
    angle = 90,
    code = 3,
    length = 0.035,
    col = cols[as.character(U_i)]
  )
  pred_i <- data.frame(n = seq(min(n_grid), max(n_grid), length.out = 100), U = U_i)
  pred_i$I_fit <- predict(fit, pred_i)
  lines(pred_i$n, pred_i$I_fit, col = cols[as.character(U_i)], lwd = 2)
}

legend(
  "bottomright",
  legend = paste("U =", U_grid),
  col = cols,
  pch = 16,
  lwd = 2,
  bty = "n"
)
```

## Grid and Residuals

```{r grid-and-residuals, fig.width=7, fig.height=4.5}
grid_out <- fit$grid
grid_out$I_fit <- fit$fitted
grid_out$residual <- fit$residuals
grid_out[, c("n", "U", "replicate", "I_empirical", "I_fit", "residual", "r_eff")]

plot(
  grid_out$I_fit,
  grid_out$residual,
  pch = 16,
  col = cols[as.character(grid_out$U)],
  xlab = "Fitted I(n,U)",
  ylab = "Residual",
  main = "Joint fit residuals"
)
abline(h = 0, lty = 2, col = "grey40")
legend(
  "topright",
  legend = paste("U =", U_grid),
  col = cols,
  pch = 16,
  bty = "n"
)
```
