---
title: "Gaussian Spike Model Fit With GSE123025"
output: html_document
---

This tutorial checks the Gaussian spike model fit on the package GSE123025
example. The data object keeps all 1,922 cells and selects 2,000 highly variable
genes from the original 23,429-gene matrix, giving a useful spectrum for the RMT
bulk and spike separation while keeping the package data modest.

```{r setup}
library(scScale)
data(gse123025_myeloid_hvg)

counts <- gse123025_myeloid_hvg$counts
dim(counts)
```

## Fit The Spectrum

We call the pieces directly: normalize raw counts, compute covariance
eigenvalues `lambda_k = s_k^2 / n`, fit the MP bulk, stop when the bulk
log-quantile RMSE no longer improves meaningfully, and then convert the final
outlying eigenvalues to spike strengths.

```{r spike-model-fit}
prepared <- scscale_prepare_matrix(
  counts,
  input = "counts",
  target_depth = 1e4,
  center = TRUE,
  scale = TRUE
)
spectrum <- scscale_spectrum(
  prepared$matrix,
  input = "normalized",
  n = prepared$n,
  p = prepared$p
)
bulk_fit <- scscale_fit_mp_iterative(
  spectrum$lambda,
  c_X = prepared$p / prepared$n,
  max_iter = 20,
  grid_n = 4000,
  stop_metric = "qq_rmse_log",
  stop_tol = 1e-4,
  min_iter = 2
)
spikes <- scscale_spike_table(spectrum$lambda, bulk_fit, r = 12)

fit_dimensions <- data.frame(
  p = prepared$p,
  n = prepared$n,
  c_X = prepared$p / prepared$n,
  eigenvalues = nrow(spectrum)
)
iteration_metrics <- bulk_fit$history[, c(
  "iteration", "tau2", "c_X", "lambda_plus", "n_bulk", "n_spikes",
  "qq_rmse_log", "ks", "neg_log_likelihood"
)]
iteration_metrics$spike_fraction <- iteration_metrics$n_spikes / nrow(spectrum)
iteration_metrics$selected <- iteration_metrics$iteration == bulk_fit$selected_iteration

fit_dimensions
iteration_metrics
data.frame(
  stop_metric = bulk_fit$stop_metric,
  stop_reason = bulk_fit$stop_reason,
  selected_iteration = bulk_fit$selected_iteration,
  iterations_checked = bulk_fit$iterations,
  tau2 = bulk_fit$tau2,
  lambda_plus = bulk_fit$lambda_plus,
  n_spikes = length(bulk_fit$spike_index)
)
spikes[, c("rank", "lambda", "lambda_tilde", "is_spike", "d2_X", "q_X")]
```

## Minimal Fit Checks

The final iteration gives one MP edge and one spike set. The plots below show
the same fit three ways: the eigenvalue distribution, the log-eigenvalue
distribution, and empirical bulk quantiles against fitted MP quantiles.

```{r plot-spike-model-fit, fig.width=11, fig.height=3.6}
final <- bulk_fit$history[bulk_fit$history$iteration == bulk_fit$selected_iteration, ]
is_spike <- spectrum$lambda > final$lambda_plus
bulk_lambda <- spectrum$lambda[!is_spike]
mp_density <- getFromNamespace("scscale_mp_density", "scScale")
mp_quantile <- getFromNamespace("scscale_mp_quantile", "scScale")
mp_x <- seq(
  max(final$mp_lower, min(bulk_lambda)),
  final$lambda_plus,
  length.out = 400
)
mp_y <- mp_density(mp_x, c_X = final$c_X, tau2 = final$tau2)
log_mp_x <- log(mp_x)
log_mp_y <- mp_y * exp(log_mp_x)
bulk_sorted <- sort(bulk_lambda)
prob <- stats::ppoints(length(bulk_sorted))
emp_q <- bulk_sorted
mp_q <- vapply(
  prob,
  mp_quantile,
  numeric(1),
  c_X = final$c_X,
  tau2 = final$tau2,
  grid_n = 4000
)

old_par <- par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2.2, 1))
hist(
  bulk_lambda,
  breaks = 60,
  probability = TRUE,
  border = "white",
  col = "gray85",
  xlab = expression(lambda[k]),
  main = "Eigenvalue distribution"
)
lines(mp_x, mp_y, col = "#0072B2", lwd = 2)
abline(v = final$lambda_plus, col = "black", lwd = 2, lty = 2)

hist(
  log(bulk_lambda),
  breaks = 60,
  probability = TRUE,
  border = "white",
  col = "gray85",
  xlab = expression(log(lambda[k])),
  main = "Log-eigenvalue distribution"
)
lines(log_mp_x, log_mp_y, col = "#0072B2", lwd = 2)
abline(v = log(final$lambda_plus), col = "black", lwd = 2, lty = 2)

plot(
  mp_q,
  emp_q,
  pch = 16,
  col = "#0072B2",
  cex = 0.35,
  xlab = "Fitted MP quantiles",
  ylab = "Empirical bulk quantiles",
  main = "Empirical bulk QQ"
)
abline(0, 1, col = "black", lwd = 2, lty = 2)
par(old_par)
```
