---
title: "UMI-Anchored Effective Scaling With GSE164378"
output: html_document
---

This tutorial fits the five-parameter effective model to RNA--ADT mutual
information measured across UMI depths. One additional parameter, \(B_0\),
then calibrates the cell-number response around the known reference point.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(scScale)
data(gse164378_3p_citeseq_hvg)

X <- gse164378_3p_citeseq_hvg$rna_counts
Y <- gse164378_3p_citeseq_hvg$adt_counts
n_reference <- ncol(X)
U_reference <- median(colSums(X))
effective_dimension <- nrow(X)
mi_rank <- 12
rates <- c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)
cell_grid <- c(500, 1000, 2000, 3000, 4000)
n_workers <- 2L
```

## Empirical UMI calibration

At every UMI depth, empirical MI is calculated with all 4,000 cells and fixed
rank-12 RNA and ADT subspaces. The effective aspect ratio uses the 2,000-feature
RNA input dimension, \(d=2000\), rather than the retained MI rank.

```{r empirical-umi, results='hide'}
umi_observed <- scscale_empirical_inu_grid(
  X,
  Y,
  n_grid = n_reference,
  sampling_rates = rates,
  r = mi_rank,
  n_replicates = 3,
  n_workers = n_workers,
  seed = 164378
)
```

## Fit the effective model

The five fitted parameters are
\[
(I_\infty,C,\theta_Y,a,b).
\]
The nonlinear optimizer varies only \((\theta_Y,a,b)\). For each optimizer
evaluation, \((I_\infty,C)\) are solved by constrained weighted least squares.
The latent signal remains linear, \(q_{\rm lin}(U)=a+bU\). Before applying the
RMT recoverability formula, a shifted softplus keeps the effective signal
strictly above the boundary:
\[
q_0(n)=\sqrt{d/n},
\]
\[
q(n,U)=q_0(n)+q_0(n)\log\!\left[1+\exp\!\left(
\frac{q_{\rm lin}(U)-q_0(n)}{q_0(n)}\right)\right],
\]
\[
\theta_X(n,U)=\frac{q(n,U)^2-d/n}{q(n,U)[q(n,U)+1]}.
\]
The floor is monotone, introduces no fitted parameter, approaches
\(\sqrt{d/n}\) for weak latent signal, and becomes linear for strong signal.

```{r fit-effective-model}
fit <- scscale_effective_fit(
  umi_observed,
  d = effective_dimension,
  n_reference = n_reference,
  U_reference = U_reference,
  weighting = "inverse_variance",
  n_starts = 10,
  seed = 164378
)

If the observed UMI range does not identify saturation, the default constraint
\(I_\infty\le 2\max I_{\rm empirical}\) prevents an arbitrary extrapolated
ceiling. Such a fit emits a warning and stores
`I_infinity_identified = FALSE`; its observed-range curve remains usable.

data.frame(
  parameter = names(fit$coefficients),
  estimate = unname(fit$coefficients),
  row.names = NULL
)
```

## Calibrate cell-number scaling

```{r empirical-cell-validation, results='hide'}
cell_observed <- scscale_empirical_inu_grid(
  X,
  Y,
  n_grid = cell_grid,
  sampling_rates = 1,
  r = mi_rank,
  n_replicates = 3,
  n_workers = n_workers,
  seed = 264378
)
fit_gamma <- scscale_effective_cell_calibrate(
  fit,
  cell_observed,
  weighting = "equal"
)
fit <- scscale_effective_cell_calibrate_B(
  fit,
  cell_observed,
  weighting = "equal"
)

summarize_mi <- function(data, x) {
  split_data <- split(data$I_empirical, data[[x]])
  data.frame(
    x = as.numeric(names(split_data)),
    I_mean = vapply(split_data, mean, numeric(1)),
    I_sd = vapply(split_data, sd, numeric(1))
  )
}

umi_points <- summarize_mi(umi_observed, "U")
cell_points <- summarize_mi(cell_observed, "n")

umi_curve <- predict(
  fit,
  n = n_reference,
  U = seq(min(umi_points$x), U_reference, length.out = 300),
  type = "all"
)
cell_curve <- predict(
  fit,
  n = seq(min(cell_grid), max(cell_grid), length.out = 300),
  U = U_reference,
  type = "all"
)

```

The reference count \(n_0=4000\) is fixed by the UMI experiment, not fitted.
The cell-number data estimate one direct scale, \(B_0\), at the reference UMI
depth \(U_0\). The original theoretical scale is
\[
A_n(U)=
\frac{\theta_Yd}
{q(U)\{q(U)+1\}\{1-\theta_Yq(U)/(q(U)+1)\}},
\]
and its relative UMI dependence transfers the fitted scale:
\[
B(U)=B_0\frac{A_n(U)}{A_n(U_0)}.
\]
Let \(I_0(U)\) be the UMI-fit prediction at \(n_0\), translated by the single
empirical anchor at \((n_0,U_0)\). The cell-number prediction is
\[
I_B(n,U)=I_0(U)+C
\log\left\{
\frac{1+B(U)/n_0}{1+B(U)/n}
\right\}.
\]
All five UMI-fit parameters remain fixed. Only \(B_0\) is estimated from
cell-number observations. Because only the ratio \(A_n(U)/A_n(U_0)\) is used,
the arbitrary absolute scale of \(q(U)\) does not enter the cell calibration.
No BBP indicator or additional gate is applied to this correction.

```{r cell-calibration-summary}
data.frame(
  parameter = "B_reference",
  estimate = unname(fit$coefficients[["B_reference"]]),
  anchor_n0 = fit$n_reference,
  anchor_I0 = fit$cell_anchor$I_empirical,
  theoretical_anchor = fit$cell_anchor$I_model,
  anchor_offset = fit$cell_anchor$offset,
  row.names = NULL
)
```

```{r scaling-curves, fig.width=10, fig.height=4.2, fig.cap="Solid lines are effective-model fits. Points and error bars are empirical MI means and standard deviations.", echo=FALSE}
plot_panel <- function(curve_x, curve_y, points, xlab, title) {
  ylim <- range(
    curve_y,
    points$I_mean - points$I_sd,
    points$I_mean + points$I_sd,
    finite = TRUE
  )
  plot(
    curve_x,
    curve_y,
    type = "l",
    lwd = 2,
    col = "#0072B2",
    xlab = xlab,
    ylab = "RNA-ADT empirical MI",
    main = title,
    ylim = ylim
  )
  segments(
    points$x,
    points$I_mean - points$I_sd,
    points$x,
    points$I_mean + points$I_sd,
    col = "grey55"
  )
  points(points$x, points$I_mean, pch = 16, col = "#222222")
}

old_par <- par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.3, 0.8))
plot_panel(
  umi_curve$U,
  umi_curve$I_fitted,
  umi_points,
  "Median RNA UMI per cell",
  "UMI scaling"
)
plot_panel(
  cell_curve$n,
  cell_curve$I_fitted,
  cell_points,
  "Cell number",
  "Cell-number scaling"
)
par(old_par)
```

```{r validation-summary}
data.frame(
  cell_model = c("Previous gamma calibration", "Direct B0 calibration"),
  parameter = c("gamma", "B_reference"),
  estimate = c(
    unname(fit_gamma$coefficients[["gamma"]]),
    unname(fit$coefficients[["B_reference"]])
  ),
  UMI_fit_RMSE = fit$metrics$rmse,
  cell_fit_RMSE = c(fit_gamma$cell_metrics$rmse, fit$cell_metrics$rmse)
)
```

## Joint cell-number and UMI scaling

The fitted model directly defines \(I(n,U)\). The curves below vary cell number
at five observed UMI depths; no additional parameters are fitted.

```{r joint-scaling, fig.width=9, fig.height=5.2, fig.cap="Predicted cell-number scaling at five observed median RNA UMI depths.", echo=FALSE}
joint_n <- seq(min(cell_grid), max(cell_grid), length.out = 300)
joint_indices <- unique(round(seq(1, nrow(umi_points), length.out = 5)))
joint_U <- umi_points$x[joint_indices]
joint_colors <- hcl.colors(length(joint_U), "Dark 3")

joint_curves <- lapply(joint_U, function(U_level) {
  predict(fit, n = joint_n, U = U_level, type = "all")
})
joint_ylim <- range(vapply(
  joint_curves,
  function(curve) range(curve$I_fitted),
  numeric(2)
))

old_par <- par(mar = c(5.1, 4.5, 4.1, 8), xpd = NA)
plot(
  joint_curves[[1]]$n,
  joint_curves[[1]]$I_fitted,
  type = "l",
  lwd = 2,
  col = joint_colors[[1]],
  xlab = "Cell number",
  ylab = expression(I(n, U)),
  main = "Joint cell-number and UMI scaling",
  ylim = joint_ylim
)
for (i in seq_along(joint_curves)[-1]) {
  lines(
    joint_curves[[i]]$n,
    joint_curves[[i]]$I_fitted,
    lwd = 2,
    col = joint_colors[[i]]
  )
}
legend(
  "right",
  inset = c(-0.25, 0),
  legend = format(round(joint_U), big.mark = ","),
  title = "Median UMI/cell",
  col = joint_colors,
  lwd = 2,
  bty = "n"
)
par(old_par)
```
