---
title: "Direct Empirical Scaling Fit With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial runs the direct empirical scaling workflow on the same
GSE164378 3' CITE-seq RNA-ADT example used by the Gaussian spike tutorials.
Here we do not fit the spike model. We sample a grid of cell numbers and UMI
depths, compute fixed-rank empirical RNA-ADT mutual information, and fit the
empirical scaling law directly.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

rna_counts <- gse164378_3p_citeseq_hvg$rna_counts
adt_counts <- gse164378_3p_citeseq_hvg$adt_counts

available_cores <- parallel::detectCores(logical = TRUE)
n_workers <- if (is.na(available_cores)) 2L else max(1L, min(4L, available_cores))
```

## Fit Empirical Scaling Law

```{r fit-empirical-scaling}
fit <- scscale_empirical_scaling_fit(
  rna_counts,
  adt_counts,
  r = 12,
  joint_mode = "hybrid",
  seed = 164378,
  count_transform = "log1p_cpm",
  n_workers = n_workers
)

fit
```

## Inspect The Fit

```{r inspect-fit, results='hide'}
empirical_mi <- scscale_empirical_values(fit)
empirical_mi_summary <- scscale_empirical_values(fit, summarize = TRUE)
parameters <- scscale_empirical_parameters(fit)
```

## Predict Scaling Laws

```{r predict-scaling-laws, results='hide'}
cell_curve <- scscale_predict_empirical_cell(
  fit,
  cell_sampling_rate = seq(0.125, 1, length.out = 200)
)

umi_curve <- scscale_predict_empirical_umi(
  fit,
  sampling_rate = seq(0.10, 1, length.out = 200)
)

joint_curve <- scscale_predict_empirical_joint(
  fit,
  cell_sampling_rate = seq(0.125, 1, length.out = 200),
  sampling_rate = c(0.20, 0.50, 1.00)
)
```

## Marginal Scaling Fits

```{r marginal-scaling-fits, fig.width=11, fig.height=4.5}
scscale_plot_empirical_fit(fit)
```

## Joint Scaling By UMI Level

```{r joint-scaling-by-umi, fig.width=8, fig.height=5.2}
scscale_plot_empirical_umi_levels(fit, sampling_rate = c(0.20, 0.50, 1.00))
```

## Joint Scaling Extrapolation

```{r joint-scaling-extrapolation, fig.width=8, fig.height=5.2}
extrapolated_joint <- scscale_predict_empirical_joint(
  fit,
  cell_sampling_rate = seq(1e-3, 2.00, length.out = 400),
  sampling_rate = c(0.20, 0.50, 1.00, 2.00, 4.00)
)

rho_levels <- sort(unique(extrapolated_joint$sampling_rate))
rho_cols <- setNames(grDevices::hcl.colors(length(rho_levels), "Dark 3"), rho_levels)

plot(
  NA,
  xlim = c(0, 2),
  ylim = range(extrapolated_joint$I_fit),
  xlab = expression("Cell sampling rate " * gamma),
  ylab = expression(hat(I)(n,U)),
  main = "Joint scaling extrapolation"
)

for (rho in rho_levels) {
  curve_i <- extrapolated_joint[extrapolated_joint$sampling_rate == rho, ]
  lines(
    curve_i$cell_sampling_rate,
    curve_i$I_fit,
    col = rho_cols[as.character(rho)],
    lwd = 2
  )
}

legend(
  "bottomright",
  legend = paste0("rho = ", rho_levels),
  col = rho_cols[as.character(rho_levels)],
  lwd = 2,
  bty = "n"
)
```
