---
title: "Complete RNA-ADT Scaling With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial fits the complete `scScale` model to 4,000 matched GSE164378
3' CITE-seq PBMC cells. The RNA matrix has 2,000 highly variable genes and the
ADT matrix has 228 antibody features.

The main interface takes the two matched feature-by-cell count matrices and
returns the full-depth fit, UMI scaling, cell-number scaling, and joint
\(I(n,U)\) grid in one object. We then estimate the corresponding empirical
MI grid directly from downsampled cells and counts.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

X <- gse164378_3p_citeseq_hvg$rna_counts
Y <- gse164378_3p_citeseq_hvg$adt_counts
```

## Fit Everything

The default UMI grid is
`c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)`. We use four absolute cell
numbers so the direct empirical comparison remains quick enough for a
tutorial.

```{r complete-fit}
n_grid <- c(500, 1000, 2000, 4000)

model <- scscale(
  X,
  Y,
  n_grid = n_grid,
  seed = 164378
)
```

The fit summary is deliberately compact:

```{r inspect-fit}
data.frame(
  cells = model$pair$x_fit$n,
  RNA_features = model$pair$x_fit$p,
  ADT_features = model$pair$y_fit$p,
  RNA_spikes = model$pair$r_X,
  ADT_spikes = model$pair$r_Y
)
```

The full results remain available in `model$pair`, `model$umi`, `model$cell`,
and `model$joint`. Per-spike linear UMI parameters are stored in
`model$parameters$q_fit_rate`.

## Empirical Comparison

`scscale_empirical_inu_grid()` performs the direct comparison. At each
\((n,U)\) grid point it samples cells, downsamples RNA counts, and measures the
RNA-ADT subspace overlap without using the spike-model prediction.

```{r empirical-grid}
empirical_grid <- scscale_empirical_inu_grid(
  X,
  Y,
  n_grid = n_grid,
  sampling_rates = model$grids$umi_sampling_rates,
  r = 12,
  n_replicates = 2,
  n_workers = 2,
  seed = 264378
)

empirical_grid$n_requested <- round(
  empirical_grid$cell_sampling_rate * ncol(X)
)

empirical_mean <- aggregate(
  I_empirical ~ sampling_rate + n_requested,
  data = empirical_grid,
  FUN = mean
)
empirical_sd <- aggregate(
  I_empirical ~ sampling_rate + n_requested,
  data = empirical_grid,
  FUN = stats::sd
)
names(empirical_sd)[names(empirical_sd) == "I_empirical"] <- "I_empirical_sd"
empirical <- merge(
  empirical_mean,
  empirical_sd,
  by = c("sampling_rate", "n_requested"),
  sort = TRUE
)
```

## Theory and Empirical Results Side by Side

Each row shows the same scaling experiment twice: spike-model theory on the
left and direct empirical subspace MI on the right. Corresponding panels use
the same y-axis range.

```{r comparison-plots, fig.width=12, fig.height=12}
draw_error_bars <- function(x, y, sd, col, length = 0.04) {
  keep <- is.finite(x) & is.finite(y) & is.finite(sd) & sd > 1e-8
  if (!any(keep)) return(invisible(NULL))
  arrows(
    x[keep],
    y[keep] - sd[keep],
    x[keep],
    y[keep] + sd[keep],
    angle = 90,
    code = 3,
    length = length,
    col = col
  )
}

rates <- sort(model$grids$umi_sampling_rates)
rate_cols <- setNames(
  grDevices::hcl.colors(length(rates), palette = "Dark 3"),
  as.character(rates)
)

umi_empirical <- empirical[
  empirical$n_requested == max(n_grid),
  ,
  drop = FALSE
]
cell_empirical <- empirical[
  empirical$sampling_rate == 1,
  ,
  drop = FALSE
]

umi_ylim <- range(
  model$umi$curve$I_theory,
  model$umi$curve$I_infinity,
  umi_empirical$I_empirical,
  finite = TRUE
)
cell_ylim <- range(
  model$cell$I_theory,
  model$pair$I_infinity,
  cell_empirical$I_empirical,
  finite = TRUE
)
joint_ylim <- range(
  model$joint$I_theory,
  empirical$I_empirical,
  finite = TRUE
)

old_par <- par(mfrow = c(3, 2), mar = c(4.2, 4.3, 2.5, 1))

plot(
  model,
  type = "umi",
  ylim = umi_ylim,
  main = "UMI scaling: theory"
)

plot(
  umi_empirical$sampling_rate,
  umi_empirical$I_empirical,
  type = "b",
  pch = 17,
  col = "#D55E00",
  ylim = umi_ylim,
  xlab = expression("UMI sampling rate " * rho),
  ylab = expression(I[empirical]),
  main = "UMI scaling: empirical"
)
draw_error_bars(
  umi_empirical$sampling_rate,
  umi_empirical$I_empirical,
  umi_empirical$I_empirical_sd,
  col = "#D55E00"
)

plot(
  model,
  type = "cell",
  ylim = cell_ylim,
  main = "Cell-number scaling: theory"
)

plot(
  cell_empirical$n_requested,
  cell_empirical$I_empirical,
  type = "b",
  pch = 17,
  col = "#D55E00",
  ylim = cell_ylim,
  xlab = "Cell number n",
  ylab = expression(I[empirical]),
  main = "Cell-number scaling: empirical"
)
draw_error_bars(
  cell_empirical$n_requested,
  cell_empirical$I_empirical,
  cell_empirical$I_empirical_sd,
  col = "#D55E00"
)

plot(
  model,
  type = "joint",
  ylim = joint_ylim,
  main = expression("Joint " * I(n, U) * ": theory")
)

plot(
  NA_real_,
  xlim = range(n_grid),
  ylim = joint_ylim,
  xlab = "Cell number n",
  ylab = expression(I[empirical](n, rho)),
  main = expression("Joint " * I(n, U) * ": empirical")
)
for (rho in rates) {
  curve <- empirical[
    empirical$sampling_rate == rho,
    ,
    drop = FALSE
  ]
  curve <- curve[order(curve$n_requested), , drop = FALSE]
  col <- rate_cols[[as.character(rho)]]
  lines(curve$n_requested, curve$I_empirical, col = col, lwd = 2)
  points(curve$n_requested, curve$I_empirical, col = col, pch = 16)
  draw_error_bars(
    curve$n_requested,
    curve$I_empirical,
    curve$I_empirical_sd,
    col = col,
    length = 0.03
  )
}
legend(
  "bottomright",
  legend = paste0("rho=", rates),
  col = rate_cols,
  pch = 16,
  lwd = 2,
  bty = "n"
)

par(old_par)
```

## Optional Controls

The default call should be sufficient for the standard workflow. The sampled
UMI depths and evaluated cell counts can still be overridden explicitly:

```{r custom-grids, eval=FALSE}
model_custom <- scscale(
  X,
  Y,
  umi_sampling_rates = c(0.20, 0.40, 0.60, 0.80, 1.00),
  n_grid = c(500, 1000, 2000, 4000),
  seed = 164378
)
```
