---
title: "Complete RNA-ADT Scaling With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial fits the complete `scScale` model to 4,000 matched GSE164378
3' CITE-seq PBMC cells. The RNA matrix has 2,000 highly variable genes and the
ADT matrix has 228 antibody features.

The main interface takes the two matched feature-by-cell count matrices and
returns the full-depth fit, UMI scaling, cell-number scaling, and joint
\(I(n,U)\) grid in one object.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

X <- gse164378_3p_citeseq_hvg$rna_counts
Y <- gse164378_3p_citeseq_hvg$adt_counts

dim(X)
dim(Y)
```

## Fit Everything

The default UMI grid is
`c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)`. The cell-number grid is
generated automatically using absolute, log-spaced cell counts.

```{r complete-fit}
model <- scscale(
  X,
  Y,
  seed = 164378
)

model
```

The complete result is organized into four main components:

```{r inspect-fit}
names(model)

names(model$pair)       # Full-depth spike fits and RNA-ADT alignment
model$umi$curve         # UMI-depth scaling
model$cell              # Cell-number scaling
model$joint             # Joint I(n,U) grid
```

The fitted linear UMI parameters for every full-depth RNA spike are saved
directly in the model:

```{r umi-parameters}
model$parameters$q_fit_rate
model$parameters$q_total_rate_fit
```

## UMI Scaling

```{r plot-umi, fig.width=7, fig.height=5}
plot(model, type = "umi")
```

## Cell-Number Scaling

```{r plot-cell, fig.width=7, fig.height=5}
plot(model, type = "cell")
```

## Joint Cell-Number by UMI Scaling

Each colored line is one UMI sampling rate evaluated over the same absolute
cell-number grid.

```{r plot-joint, fig.width=8, fig.height=5.5}
plot(model, type = "joint")
```

## Optional Controls

The default call should be sufficient for the standard workflow. The sampled
UMI depths and evaluated cell counts can still be overridden explicitly:

```{r custom-grids, eval=FALSE}
model_custom <- scscale(
  X,
  Y,
  umi_sampling_rates = c(0.20, 0.40, 0.60, 0.80, 1.00),
  n_grid = c(500, 1000, 2000, 4000),
  seed = 164378
)
```
