---
title: "Sequencing-Depth Scaling With GSE123025"
output: html_document
---

This tutorial tests the linear sequencing-depth scaling model for fitted spike
strengths. Each sequencing condition is a global molecule-thinning experiment:
for sampling rate `rho`, every count is sampled with probability `rho`, all
cells are retained, and the x-axis is the expected total UMI count
`rho * sum(counts)`.

```{r setup}
library(scScale)
data(gse123025_myeloid_hvg)

counts <- gse123025_myeloid_hvg$counts
dim(counts)
summary(colSums(counts))
```

## Fit UMI Scaling

The fit uses full-depth feature directions as the reference, then projects each
downsampled matrix onto those directions. For each spike `k`, we fit
`q_X,k(U) = alpha_k + beta_k U`.

```{r fit-umi-scaling}
sampling_rates <- c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)

umi_scaling <- scscale_umi_scaling(
  counts,
  sampling_rates = sampling_rates,
  r = 12,
  seed = 11,
  n_replicates = 1,
  empirical = FALSE
)

unique(umi_scaling$scaling[, c("sampling_rate", "replicate", "U", "total_umi_observed", "p", "n")])
umi_scaling$q_fit
umi_scaling$q_fit_rate
umi_scaling$q_total_fit
umi_scaling$q_total_rate_fit
```

## Linear Fits

```{r plot-umi-linear-fits, fig.width=11, fig.height=3.7}
top_ranks <- 1:6
plot_q <- subset(umi_scaling$scaling, rank %in% top_ranks)
plot_total <- subset(umi_scaling$q_total, sampling_rate > 0.1)
plot_total_model <- lm(q_total ~ sampling_rate, data = plot_total)
plot_total$q_total_rate_hat <- predict(plot_total_model, newdata = plot_total)
plot_total_r2 <- summary(plot_total_model)$r.squared
cols <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#56B4E9", "#E69F00")

old_par <- par(mfrow = c(1, 3), mar = c(4.2, 4.2, 2.4, 1))
plot(
  NA,
  xlim = range(plot_q$sampling_rate),
  ylim = range(c(plot_q$q_X, plot_q$q_X_rate_hat)),
  xlab = expression("Sampling rate " * rho),
  ylab = expression(q[X*k](rho)),
  main = "Top spike fits"
)
for (i in seq_along(top_ranks)) {
  k <- top_ranks[i]
  df <- plot_q[plot_q$rank == k, ]
  points(df$sampling_rate, df$q_X, pch = 16, col = cols[i])
  fit_df <- unique(df[, c("sampling_rate", "q_X_rate_hat")])
  fit_df <- fit_df[order(fit_df$sampling_rate), ]
  lines(fit_df$sampling_rate, fit_df$q_X_rate_hat, col = cols[i], lwd = 2)
}
legend(
  "topleft",
  legend = paste0("k=", top_ranks),
  col = cols,
  pch = 16,
  lwd = 2,
  bty = "n",
  cex = 0.85
)

plot(
  plot_total$sampling_rate,
  plot_total$q_total,
  pch = 16,
  col = "#0072B2",
  xlab = expression("Sampling rate " * rho),
  ylab = expression(sum(q[X*k](rho))),
  main = "Total spike strength"
)
lines(
  plot_total$sampling_rate,
  plot_total$q_total_rate_hat,
  col = "#D55E00",
  lwd = 2
)
legend(
  "topleft",
  legend = c("observed", "linear fit"),
  col = c("#0072B2", "#D55E00"),
  pch = c(16, NA),
  lwd = c(NA, 2),
  bty = "n"
)

plot(
  umi_scaling$q_fit_rate$rank,
  umi_scaling$q_fit_rate$q_rate_r2,
  type = "h",
  lwd = 4,
  col = "#0072B2",
  ylim = c(0, 1),
  xlab = "Spike rank k",
  ylab = expression(R^2),
  main = expression(R^2 * " across spikes")
)
points(umi_scaling$q_fit_rate$rank, umi_scaling$q_fit_rate$q_rate_r2, pch = 16, col = "#0072B2")
abline(h = plot_total_r2, col = "#D55E00", lwd = 2, lty = 2)
legend(
  "bottomright",
  legend = expression("total " * R^2),
  col = "#D55E00",
  lwd = 2,
  lty = 2,
  bty = "n"
)
par(old_par)
```
