---
title: "RNA-ADT Mutual Information With GSE164378 3' CITE-seq"
output: html_document
---

This tutorial runs the paired-modality `scScale` workflow on 4,000 matched
GSE164378 3' CITE-seq PBMC cells. The RNA side has 2,000 HVGs and the ADT side
has 228 antibody features.

The package-level scaling API has three pieces:

1. `scscale_umi_mi()` for UMI-depth scaling.
2. `scscale_cell_number_mi()` for cell-number scaling.
3. `scscale_cell_number_by_umi_mi()` for the joint cell-number by UMI grid.

Empirical curves are direct calls to `scscale_empirical_mi()`.
All three theory functions use the same low-rank MI formula,
`scscale_low_rank_mi()`, after preparing the appropriate recoverability values
and RNA-ADT subspace alignment matrix.

```{r setup}
library(scScale)
data(gse164378_3p_citeseq_hvg)

rna_counts <- gse164378_3p_citeseq_hvg$rna_counts
adt_counts <- gse164378_3p_citeseq_hvg$adt_counts
count_transform <- "log1p_cpm"
r_empirical <- 12

dim(rna_counts)
dim(adt_counts)
```

## Fit Modalities

`scscale_pair_fit()` fits MAD-based spike models for RNA and ADT, then builds
the RNA-ADT subspace-overlap matrix used by theoretical MI.

```{r fit-modalities}
pair <- scscale_pair_fit(
  rna_counts,
  adt_counts,
  count_transform = count_transform,
  mp_grid_n = 3000
)

data.frame(
  modality = c("RNA", "ADT"),
  p = c(pair$x_fit$p, pair$y_fit$p),
  n = c(pair$x_fit$n, pair$y_fit$n),
  n_spikes = c(pair$r_X, pair$r_Y),
  tau2 = c(pair$x_fit$bulk$tau2, pair$y_fit$bulk$tau2),
  lambda_plus = c(pair$x_fit$bulk$lambda_plus, pair$y_fit$bulk$lambda_plus),
  method = c(pair$x_fit$bulk$method, pair$y_fit$bulk$method)
)

data.frame(
  I_theory = pair$mi$I_theory,
  I_empirical = pair$mi$I_empirical,
  I_infinity = pair$I_infinity
)
```

## Theory Curves

The UMI curve refits the RNA spike model at each depth, so both
`I_theory` and `I_infinity` use the BBP-detectable rank at that UMI level.

```{r theory-curves}
sampling_rates <- c(0.10, 0.20, 0.35, 0.50, 0.70, 0.85, 1.00)
rho_grid <- c(0.35, 0.50, 0.70, 0.85, 1.00)
n_grid <- c(500, 1000, 2000, 4000)

umi_theory <- scscale_umi_mi(
  pair,
  rna_counts,
  sampling_rates = sampling_rates,
  count_transform = count_transform,
  seed = 164378
)

cell_theory <- scscale_cell_number_mi(pair, n_grid = n_grid)

cell_by_umi_theory <- scscale_cell_number_by_umi_mi(
  pair,
  umi_theory,
  n_grid = n_grid,
  sampling_rates = rho_grid
)

umi_theory$curve[, c("sampling_rate", "r_X", "I_theory", "I_infinity")]
```

## Empirical Curves

The empirical curves use the same subspace-overlap MI estimator directly on
downsampled or subsetted count matrices.

```{r empirical-curves}
summarize_empirical <- function(x) {
  out <- aggregate(I_empirical ~ sampling_rate + n, data = x, FUN = mean)
  out$I_empirical_sd <- aggregate(I_empirical ~ sampling_rate + n, data = x, FUN = stats::sd)$I_empirical
  out$n_replicates <- length(unique(x$replicate))
  out
}

umi_empirical <- do.call(rbind, lapply(seq_along(sampling_rates), function(i) {
  rho <- sampling_rates[i]
  rna_sub <- rna_counts
  if (rho < 1) {
    rna_sub <- scscale_downsample_counts_fraction(rna_sub, fraction = rho, seed = 364378 + i)
  }
  mi <- scscale_empirical_mi(rna_sub, adt_counts, r = r_empirical, count_transform = count_transform)
  data.frame(sampling_rate = rho, I_empirical = mi$mi, r_eff = mi$r_eff)
}))

n_replicates <- 2
cell_orders <- lapply(seq_len(n_replicates), function(replicate) {
  set.seed(264378 + replicate)
  sample(colnames(rna_counts), max(n_grid))
})

cell_empirical_replicates <- do.call(rbind, lapply(seq_len(n_replicates), function(replicate) {
  do.call(rbind, lapply(n_grid, function(n_i) {
    cells <- cell_orders[[replicate]][seq_len(n_i)]
    mi <- scscale_empirical_mi(
      rna_counts[, cells, drop = FALSE],
      adt_counts[, cells, drop = FALSE],
      r = r_empirical,
      count_transform = count_transform
    )
    data.frame(sampling_rate = 1, n = n_i, replicate = replicate, I_empirical = mi$mi, r_eff = mi$r_eff)
  }))
}))
cell_empirical <- summarize_empirical(cell_empirical_replicates)

cell_by_umi_empirical_replicates <- do.call(rbind, lapply(seq_along(rho_grid), function(rho_i) {
  rho <- rho_grid[rho_i]
  do.call(rbind, lapply(seq_len(n_replicates), function(replicate) {
    do.call(rbind, lapply(n_grid, function(n_i) {
      cells <- cell_orders[[replicate]][seq_len(n_i)]
      rna_sub <- rna_counts[, cells, drop = FALSE]
      if (rho < 1) {
        rna_sub <- scscale_downsample_counts_fraction(
          rna_sub,
          fraction = rho,
          seed = 464378 + rho_i * 100 + replicate * 10 + match(n_i, n_grid)
        )
      }
      mi <- scscale_empirical_mi(
        rna_sub,
        adt_counts[, cells, drop = FALSE],
        r = r_empirical,
        count_transform = count_transform
      )
      data.frame(sampling_rate = rho, n = n_i, replicate = replicate, I_empirical = mi$mi, r_eff = mi$r_eff)
    }))
  }))
}))
cell_by_umi_empirical <- summarize_empirical(cell_by_umi_empirical_replicates)
```

## Six Plots

```{r six-panel-plots, fig.width=12, fig.height=12}
draw_error_bars <- function(x, y, sd, col, length = 0.04) {
  keep <- is.finite(x) & is.finite(y) & is.finite(sd) & sd > 1e-8
  if (!any(keep)) return(invisible(NULL))
  arrows(x[keep], y[keep] - sd[keep], x[keep], y[keep] + sd[keep],
         angle = 90, code = 3, length = length, col = col)
}

rho_cols <- setNames(
  c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#E69F00"),
  as.character(rho_grid)
)

old_par <- par(mfrow = c(3, 2), mar = c(4.2, 4.3, 2.5, 1))

plot(
  umi_theory$curve$sampling_rate,
  umi_theory$curve$I_theory,
  type = "b",
  pch = 16,
  col = "#0072B2",
  ylim = range(c(umi_theory$curve$I_theory, umi_theory$curve$I_infinity)),
  xlab = expression("Sampling rate " * rho),
  ylab = expression(I[theory]),
  main = "UMI scaling: theory"
)
lines(umi_theory$curve$sampling_rate, umi_theory$curve$I_infinity, col = "grey35", lwd = 2, lty = 2)
legend("bottomright", legend = c("refit rank", expression(I[infinity])),
       col = c("#0072B2", "grey35"), pch = c(16, NA), lwd = c(1, 2),
       lty = c(1, 2), bty = "n")

plot(
  umi_empirical$sampling_rate,
  umi_empirical$I_empirical,
  type = "b",
  pch = 17,
  col = "#D55E00",
  xlab = expression("Sampling rate " * rho),
  ylab = expression(I[empirical]),
  main = "UMI scaling: empirical"
)

plot(
  cell_theory$n,
  cell_theory$I_theory,
  type = "b",
  pch = 16,
  col = "#0072B2",
  ylim = range(c(cell_theory$I_theory, pair$I_infinity)),
  xlab = "Cell number n",
  ylab = expression(I[theory]),
  main = "Cell number scaling: theory"
)
abline(h = pair$I_infinity, col = "grey35", lwd = 2, lty = 2)
legend("topright", legend = c("theory", expression(I[infinity])),
       col = c("#0072B2", "grey35"), pch = c(16, NA), lwd = c(1, 2),
       lty = c(1, 2), bty = "n")

plot(
  cell_empirical$n,
  cell_empirical$I_empirical,
  type = "b",
  pch = 17,
  col = "#D55E00",
  xlab = "Cell number n",
  ylab = expression(I[empirical]),
  main = "Cell number scaling: empirical"
)
draw_error_bars(cell_empirical$n, cell_empirical$I_empirical, cell_empirical$I_empirical_sd, col = "#D55E00")

plot(
  NA,
  xlim = range(n_grid),
  ylim = range(cell_by_umi_theory$I_theory),
  xlab = "Cell number n",
  ylab = expression(I[theory](n, rho)),
  main = "Cell number by UMI: theory"
)
for (rho in rho_grid) {
  key <- as.character(rho)
  df <- cell_by_umi_theory[cell_by_umi_theory$sampling_rate == rho, ]
  df <- df[order(df$n), ]
  lines(df$n, df$I_theory, col = rho_cols[key], lwd = 2)
  points(df$n, df$I_theory, col = rho_cols[key], pch = 16)
}
legend("topright", legend = paste0("rho=", rho_grid), col = rho_cols, lwd = 2, pch = 16, bty = "n")

plot(
  NA,
  xlim = range(n_grid),
  ylim = range(cell_by_umi_empirical$I_empirical),
  xlab = "Cell number n",
  ylab = expression(I[empirical](n, rho)),
  main = "Cell number by UMI: empirical"
)
for (rho in rho_grid) {
  key <- as.character(rho)
  emp <- cell_by_umi_empirical[cell_by_umi_empirical$sampling_rate == rho, ]
  emp <- emp[order(emp$n), ]
  lines(emp$n, emp$I_empirical, col = rho_cols[key], lwd = 2)
  points(emp$n, emp$I_empirical, col = rho_cols[key], pch = 16)
  draw_error_bars(emp$n, emp$I_empirical, emp$I_empirical_sd, col = rho_cols[key], length = 0.03)
}
legend("topright", legend = paste0("rho=", rho_grid), col = rho_cols, lwd = 2, pch = 16, bty = "n", cex = 0.85)

par(old_par)
```
