---
title: "Batch number scaling law"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "figure/batch-number-"
)

local_lib <- normalizePath(file.path("..", "..", "..", ".rlib"), mustWork = FALSE)
if (dir.exists(local_lib)) .libPaths(c(local_lib, .libPaths()))
library(scScale)
set.seed(17)
```

## Goal

This tutorial shows the batch-effect scaling law added from the GitHub BMMC
CITE-seq script:

`I = I_inf - C * log(1 - A / m)`

where `m` is the number of batches at fixed total cells.

The original script used paired RNA and ADT modalities. This lightweight
tutorial starts from a real Jurkat 10x RNA count block and splits its HVGs into
two disjoint gene views. That gives us paired cell-side matrices without
shipping a large multimodal object.

## Load real Jurkat counts and make two views

`batch_effect_mi()` estimates biological information from two feature-by-cell
matrices by computing cell-side subspaces and squared canonical correlations.

```{r load-counts}
counts_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_counts_hvg.csv"
)

counts <- as.matrix(read.csv(counts_file, row.names = 1, check.names = FALSE))
dim(counts)

view_a <- counts[seq_len(200), , drop = FALSE]
view_b <- counts[201:400, , drop = FALSE]
```

```{r one-mi}
cells_demo <- sample(colnames(counts), 600)

bio <- batch_effect_mi(
  view_a[, cells_demo, drop = FALSE],
  view_b[, cells_demo, drop = FALSE],
  r_x = 5,
  r_y = 5,
  transform_x = "log1p",
  transform_y = "log1p"
)

bio[c("mi", "mi_norm", "r_eff")]
```

## Sampling cells by batch

`sample_batch_cells()` samples a balanced number of cells from each selected
batch. The Jurkat block does not come with experimental batch labels, so we
create pseudo-batches by ordering cells by UMI depth and cutting them into
equally sized groups. Real analyses should use an experimental batch or donor
column.

```{r sample-batches}
cell_depth <- colSums(counts)
batch_id <- cut(
  rank(cell_depth, ties.method = "first"),
  breaks = 12,
  labels = paste0("depth_batch_", seq_len(12))
)

meta <- data.frame(
  batch = as.character(batch_id),
  umi_in_hvg_block = cell_depth,
  row.names = colnames(counts)
)

sampled <- sample_batch_cells(
  meta = meta,
  batch_col = "batch",
  m_batch = 4,
  cells_per_batch = 80,
  seed = 1
)

head(sampled$cells)
sampled$batches
```

## Fit the batch-number law

For each design point below, we sample cells from a fixed total of 480 cells
and vary the number of batches.

```{r make-replicates}
run_design <- function(m_batch, rep_id, n_total = 480) {
  cells_per_batch <- as.integer(n_total / m_batch)
  sampled <- sample_batch_cells(
    meta = meta,
    batch_col = "batch",
    m_batch = m_batch,
    cells_per_batch = cells_per_batch,
    seed = 1000 + 100 * rep_id + m_batch
  )
  bio <- batch_effect_mi(
    view_a[, sampled$cells, drop = FALSE],
    view_b[, sampled$cells, drop = FALSE],
    r_x = 5,
    r_y = 5,
    transform_x = "log1p",
    transform_y = "log1p"
  )
  data.frame(
    experiment = "fixed_n_vary_m",
    rep = rep_id,
    m_batch = m_batch,
    cells_per_batch = cells_per_batch,
    n_cells = length(sampled$cells),
    I_bio = bio$mi,
    I_bio_norm = bio$mi_norm
  )
}

replicate_results <- do.call(
  rbind,
  lapply(seq_len(3), function(rep_id) {
    do.call(rbind, lapply(c(2, 3, 4, 6, 8, 12), run_design, rep_id = rep_id))
  })
)

summary_df <- summarize_batch_effect_results(replicate_results)
summary_df
```

```{r fit-batch-law}
batch_fit <- suppressWarnings(
  fit_batch_scaling(
    summary_df,
    law = "batch_number",
    target_col = "mean_I_bio_norm",
    min_points = 5
  )
)

coef(batch_fit)
summary(batch_fit)
predict(batch_fit, data.frame(m_batch = c(16, 24, 32)))
```

## Plot

```{r plot-batch-law}
plot(batch_fit, xlab = "Number of pseudo-batches", ylab = "Mean normalized biological information")
```

## Cells per batch variant

The companion law uses fixed batch number and varies cells per batch:

`I = I_inf - C * log(1 + A / s)`

Use the same function with `law = "cells_per_batch"` and a
`cells_per_batch` column in the summary table.
