---
title: "Batch number scaling law"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "figure/batch-number-"
)

local_lib <- normalizePath(file.path("..", "..", "..", ".rlib"), mustWork = FALSE)
if (dir.exists(local_lib)) .libPaths(c(local_lib, .libPaths()))
library(scScale)
set.seed(17)

make_tutorial_counts <- function(seed = 11, n_genes = 400, n_cells = 5000, n_factors = 6) {
  set.seed(seed)
  gene_loadings <- matrix(rnorm(n_genes * n_factors, sd = 0.45), nrow = n_genes)
  cell_scores <- matrix(rnorm(n_cells * n_factors), nrow = n_cells)
  gene_baseline <- rnorm(n_genes, mean = log(1.5), sd = 0.35)
  log_mu <- gene_baseline + gene_loadings %*% t(cell_scores) / sqrt(n_factors)
  counts <- matrix(rpois(length(log_mu), lambda = pmin(exp(log_mu), 80)), nrow = n_genes)
  rownames(counts) <- paste0("gene_", seq_len(n_genes))
  colnames(counts) <- paste0("cell_", seq_len(n_cells))
  counts
}
```

## Goal

This tutorial shows the batch-effect scaling law added from the GitHub BMMC
CITE-seq script:

`I = I_inf - C * log(1 - A / m)`

where `m` is the number of batches at fixed total cells.

The original script used paired RNA and ADT modalities. This lightweight
tutorial starts from the same compact count block as the other tutorials and
splits its features into two disjoint gene views. If the repository-local Jurkat
10x block is available, the tutorial uses it; otherwise it generates a small
structured count matrix so the page remains self-contained.

## Load real Jurkat counts and make two views

`batch_effect_mi()` estimates biological information from two feature-by-cell
matrices by computing cell-side subspaces and squared canonical correlations.

```{r load-counts}
counts_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_counts_hvg.csv"
)

if (file.exists(counts_file)) {
  counts <- as.matrix(read.csv(counts_file, row.names = 1, check.names = FALSE))
  counts_source <- "Jurkat 10x HVG block"
} else {
  counts <- make_tutorial_counts(seed = 11)
  counts_source <- "synthetic structured tutorial block"
}

counts_source
dim(counts)

view_a <- counts[seq_len(200), , drop = FALSE]
view_b <- counts[201:400, , drop = FALSE]
```

```{r one-mi}
cells_demo <- sample(colnames(counts), 600)

bio <- batch_effect_mi(
  view_a[, cells_demo, drop = FALSE],
  view_b[, cells_demo, drop = FALSE],
  r_x = 5,
  r_y = 5,
  transform_x = "log1p",
  transform_y = "log1p"
)

bio[c("mi", "mi_norm", "r_eff")]
```

The raw MI can be mapped to a bounded average-overlap scale:

```{r normalized-overlap-helper}
normalized_overlap <- function(mi, r) {
  1 - exp(-2 * mi / pmax(r, 1))
}

normalized_overlap(bio$mi, bio$r_eff)
```

## Sampling cells by batch

`sample_batch_cells()` samples a balanced number of cells from each selected
batch. The Jurkat block does not come with experimental batch labels, so we
create pseudo-batches by ordering cells by UMI depth and cutting them into
equally sized groups. Real analyses should use an experimental batch or donor
column.

```{r sample-batches}
cell_depth <- colSums(counts)
batch_id <- cut(
  rank(cell_depth, ties.method = "first"),
  breaks = 12,
  labels = paste0("depth_batch_", seq_len(12))
)

meta <- data.frame(
  batch = as.character(batch_id),
  umi_in_hvg_block = cell_depth,
  row.names = colnames(counts)
)

sampled <- sample_batch_cells(
  meta = meta,
  batch_col = "batch",
  m_batch = 4,
  cells_per_batch = 80,
  seed = 1
)

head(sampled$cells)
sampled$batches
```

## Fit the batch-number law

For each design point below, we sample cells from a fixed total of 480 cells
and vary the number of batches.

```{r make-replicates}
run_design <- function(m_batch, rep_id, n_total = 480) {
  cells_per_batch <- as.integer(n_total / m_batch)
  sampled <- sample_batch_cells(
    meta = meta,
    batch_col = "batch",
    m_batch = m_batch,
    cells_per_batch = cells_per_batch,
    seed = 1000 + 100 * rep_id + m_batch
  )
  bio <- batch_effect_mi(
    view_a[, sampled$cells, drop = FALSE],
    view_b[, sampled$cells, drop = FALSE],
    r_x = 5,
    r_y = 5,
    transform_x = "log1p",
    transform_y = "log1p"
  )
  data.frame(
    experiment = "fixed_n_vary_m",
    rep = rep_id,
    m_batch = m_batch,
    cells_per_batch = cells_per_batch,
    n_cells = length(sampled$cells),
    I_bio = bio$mi,
    I_bio_norm = bio$mi_norm,
    r_eff = bio$r_eff,
    I_bio_overlap = normalized_overlap(bio$mi, bio$r_eff)
  )
}

replicate_results <- do.call(
  rbind,
  lapply(seq_len(3), function(rep_id) {
    do.call(rbind, lapply(c(2, 3, 4, 6, 8, 12), run_design, rep_id = rep_id))
  })
)

summary_df <- summarize_batch_effect_results(replicate_results)
overlap_summary <- aggregate(
  cbind(r_eff, I_bio_overlap) ~ experiment + m_batch + cells_per_batch + n_cells,
  data = replicate_results,
  FUN = mean
)
names(overlap_summary)[names(overlap_summary) == "r_eff"] <- "mean_r_eff"
names(overlap_summary)[names(overlap_summary) == "I_bio_overlap"] <- "mean_I_bio_overlap"
summary_df <- merge(
  summary_df,
  overlap_summary,
  by = c("experiment", "m_batch", "cells_per_batch", "n_cells"),
  sort = FALSE
)
summary_df
```

```{r fit-batch-law}
batch_fit <- suppressWarnings(
  fit_batch_scaling(
    summary_df,
    law = "batch_number",
    target_col = "mean_I_bio_norm",
    min_points = 5
  )
)

coef(batch_fit)
summary(batch_fit)
predict(batch_fit, data.frame(m_batch = c(16, 24, 32)))
```

The same batch-number law can also be fit to the bounded normalized-overlap
score.

```{r fit-batch-overlap-law}
batch_overlap_fit <- suppressWarnings(
  fit_batch_scaling(
    summary_df,
    law = "batch_number",
    target_col = "mean_I_bio_overlap",
    min_points = 5
  )
)

coef(batch_overlap_fit)
summary(batch_overlap_fit)
predict(batch_overlap_fit, data.frame(m_batch = c(16, 24, 32)))
```

## Plot biological information

```{r plot-batch-law}
plot(batch_fit, xlab = "Number of pseudo-batches", ylab = "Mean normalized biological information")
```

## Plot empirical overlap

```{r plot-batch-overlap-law}
plot(batch_overlap_fit, xlab = "Number of pseudo-batches", ylab = "Empirical normalized overlap")
```

## Cells per batch variant

The companion law uses fixed batch number and varies cells per batch:

`I = I_inf - C * log(1 + A / s)`

Use the same function with `law = "cells_per_batch"` and a
`cells_per_batch` column in the summary table.
