---
title: "Cell number scaling law"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "figure/cell-number-"
)

local_lib <- normalizePath(file.path("..", "..", "..", ".rlib"), mustWork = FALSE)
if (dir.exists(local_lib)) .libPaths(c(local_lib, .libPaths()))
library(scScale)
set.seed(11)

make_tutorial_counts <- function(seed = 11, n_genes = 400, n_cells = 5000, n_factors = 6) {
  set.seed(seed)
  gene_loadings <- matrix(rnorm(n_genes * n_factors, sd = 0.45), nrow = n_genes)
  cell_scores <- matrix(rnorm(n_cells * n_factors), nrow = n_cells)
  gene_baseline <- rnorm(n_genes, mean = log(1.5), sd = 0.35)
  log_mu <- gene_baseline + gene_loadings %*% t(cell_scores) / sqrt(n_factors)
  counts <- matrix(rpois(length(log_mu), lambda = pmin(exp(log_mu), 80)), nrow = n_genes)
  rownames(counts) <- paste0("gene_", seq_len(n_genes))
  colnames(counts) <- paste0("cell_", seq_len(n_cells))
  counts
}
```

## Goal

This example asks how a spectral signal estimate changes as the number of cells
increases. It uses a compact 400-feature by 5,000-cell tutorial count block. If
the repository-local Jurkat 10x block is available, the tutorial uses it;
otherwise it generates a small structured count matrix so the page remains
self-contained.

The workflow is:

1. load or generate the feature-by-cell count matrix;
2. sample different numbers of cells;
3. call `fit_cell_scaling()` to fit spectra across the grid;
4. compare theoretical and empirical normalized overlap scores;
5. use `predict()` to extrapolate to larger cell counts.

## Load real Jurkat counts

```{r load-counts}
counts_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_counts_hvg.csv"
)

if (file.exists(counts_file)) {
  counts <- as.matrix(read.csv(counts_file, row.names = 1, check.names = FALSE))
  counts_source <- "Jurkat 10x HVG block"
} else {
  counts <- make_tutorial_counts(seed = 11)
  counts_source <- "synthetic structured tutorial block"
}

counts_source
dim(counts)
summary(colSums(counts))
```

The full source dataset contains many more cells when the Jurkat block is
available; this tutorial uses a compact HVG-like block so it renders quickly.

```{r source-summary}
sample_summary_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_sample_summary.csv"
)

if (file.exists(sample_summary_file)) {
  sample_summary <- read.csv(sample_summary_file)
} else {
  sample_summary <- data.frame(
    dataset = counts_source,
    n_total_cells = ncol(counts),
    sampled_cells = ncol(counts),
    n_genes_used = nrow(counts),
    mean_umi_per_cell = mean(colSums(counts))
  )
}
sample_summary[, c("dataset", "n_total_cells", "sampled_cells", "n_genes_used", "mean_umi_per_cell")]
```

## Fit the cell-number scaling object

`fit_cell_scaling()` computes the reference spectrum from the full matrix,
subsamples cells across `cell_grid`, computes theoretical MI with `mi_theory()`,
and fits a saturating curve to normalized MI.

```{r cell-fit}
cell_fit <- fit_cell_scaling(
  counts,
  cell_grid = c(250, 500, 1000, 2000, 3500, 5000),
  n_features = 300,
  transform = "log1p",
  min_cells = 10,
  r = 8,
  R = 3,
  p_sim = 100,
  seed = 1
)

cell_fit$data
```

## Empirical reference overlap

For a direct check, compare each subsampled RNA embedding with the full RNA
embedding using the same subspace-overlap MI formula. We normalize both
theoretical and empirical MI as `1 - exp(-2 * MI / r)`, which puts the scores on
the same bounded overlap scale.

```{r cell-empirical-mi}
cell_grid <- c(250, 500, 1000, 2000, 3500, 5000)

normalized_overlap <- function(mi, r) {
  1 - exp(-2 * mi / pmax(r, 1))
}

cell_embedding <- function(x, r = 8, n_features = 300, min_cells = 10) {
  features <- select_hvgs(x, n_features = n_features, min_cells = min_cells)
  right_singular_vectors(log1p(x[features, , drop = FALSE]), r = r)
}

z_full <- cell_embedding(counts, r = 8, n_features = 300, min_cells = 10)

cell_empirical <- do.call(rbind, lapply(cell_grid, function(n_cells) {
  set.seed(1 + 10000L + n_cells)
  cells <- if (n_cells == ncol(counts)) colnames(counts) else sample(colnames(counts), n_cells)
  z_sub <- cell_embedding(counts[, cells, drop = FALSE], r = 8, n_features = 300, min_cells = 10)
  emp <- subspace_overlap_mi(z_sub, z_full)
  data.frame(
    n_cells = n_cells,
    empirical_mi = emp$mi,
    empirical_r_eff = emp$r_eff,
    empirical_overlap = normalized_overlap(emp$mi, emp$r_eff)
  )
}))

cell_compare <- merge(cell_fit$data, cell_empirical, by = "n_cells")
cell_compare$theory_overlap <- normalized_overlap(cell_compare$mean_mi_norm, 1)
cell_compare[, c("n_cells", "theory_overlap", "empirical_overlap", "mean_mi_norm", "empirical_mi", "empirical_r_eff")]
```

## Inspect and extrapolate

The fitted object supports the usual R-style methods.

```{r inspect-cell-fit}
coef(cell_fit)
summary(cell_fit)
predict(cell_fit, data.frame(n_cells = c(7500, 10000, 20000)))
```

## Plot theory scaling law

```{r plot-cell-law}
plot(cell_fit, xlab = "Cells sampled", ylab = "Theoretical MI / retained rank")
```

## Plot empirical overlap

```{r plot-cell-empirical}
plot(
  cell_compare$n_cells,
  cell_compare$empirical_overlap,
  pch = 17,
  col = "#d95f02",
  type = "b",
  xlab = "Cells sampled",
  ylab = "Empirical normalized overlap"
)
```

## Adapt to real data

Replace `counts` with another real feature-by-cell matrix or Seurat object
accepted by `find_eigenvalues()`. Keep the same cell grid loop, then choose the
response that best matches the biological question.
