---
title: "UMI scaling law"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "figure/umi-"
)

local_lib <- normalizePath(file.path("..", "..", "..", ".rlib"), mustWork = FALSE)
if (dir.exists(local_lib)) .libPaths(c(local_lib, .libPaths()))
library(scScale)
set.seed(13)
```

## Goal

This example asks how a spectral signal estimate changes with sequencing depth,
represented here as total UMI depth. It starts from the same real Jurkat 10x
HVG count block used in the cell-number tutorial.

The workflow is:

1. load a real feature-by-cell count matrix;
2. depth-subsample UMIs at several fractions;
3. call `fit_umi_scaling()` to fit spectra across the depth grid;
4. inspect `coef()`, `summary()`, and `plot()`;
5. use `predict()` to extrapolate to higher depth.

## Load real Jurkat counts

```{r load-counts}
counts_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_counts_hvg.csv"
)

counts_full <- as.matrix(read.csv(counts_file, row.names = 1, check.names = FALSE))
dim(counts_full)
summary(colSums(counts_full))
```

To keep rendering fast, use 3,000 cells from the 5,000-cell tutorial block.

```{r choose-cells}
cells_use <- sample(colnames(counts_full), 3000)
counts_base <- counts_full[, cells_use, drop = FALSE]
summary(colSums(counts_base))
```

## Fit the UMI-depth scaling object

`fit_umi_scaling()` binomially thins counts to each UMI fraction, compares each
downsampled spectrum to the full-depth reference with `mi_theory()`, and fits a
saturating curve to normalized MI.

```{r umi-fit}
umi_fit <- fit_umi_scaling(
  counts_base,
  umi_grid = c(0.02, 0.05, 0.1, 0.2, 0.4, 0.7, 1.0),
  n_cells = ncol(counts_base),
  n_features = 300,
  transform = "log1p",
  min_cells = 10,
  r = 8,
  R = 3,
  p_sim = 100,
  seed = 1
)

umi_fit$data
```

## Inspect and extrapolate

```{r inspect-umi-fit}
coef(umi_fit)
summary(umi_fit)
predict(umi_fit, data.frame(mean_umi_per_cell = c(2500, 3500, 5000)))
```

## Plot

```{r plot-umi-law}
plot(umi_fit, xlab = "Mean UMI per cell in HVG block", ylab = "Theoretical MI / retained rank")
```

## Adapt to real data

Use another real count matrix as `counts_full`, repeat the binomial thinning
across fractions and replicates, and store the response you want to model. The
spectral estimation pieces stay the same.
