---
title: "UMI scaling law"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "figure/umi-"
)

local_lib <- normalizePath(file.path("..", "..", "..", ".rlib"), mustWork = FALSE)
if (dir.exists(local_lib)) .libPaths(c(local_lib, .libPaths()))
library(scScale)
set.seed(13)

make_tutorial_counts <- function(seed = 11, n_genes = 400, n_cells = 5000, n_factors = 6) {
  set.seed(seed)
  gene_loadings <- matrix(rnorm(n_genes * n_factors, sd = 0.45), nrow = n_genes)
  cell_scores <- matrix(rnorm(n_cells * n_factors), nrow = n_cells)
  gene_baseline <- rnorm(n_genes, mean = log(1.5), sd = 0.35)
  log_mu <- gene_baseline + gene_loadings %*% t(cell_scores) / sqrt(n_factors)
  counts <- matrix(rpois(length(log_mu), lambda = pmin(exp(log_mu), 80)), nrow = n_genes)
  rownames(counts) <- paste0("gene_", seq_len(n_genes))
  colnames(counts) <- paste0("cell_", seq_len(n_cells))
  counts
}
```

## Goal

This example asks how a spectral signal estimate changes with sequencing depth,
represented here as total UMI depth. It uses the same compact tutorial count
block as the cell-number tutorial. If the repository-local Jurkat 10x block is
available, the tutorial uses it; otherwise it generates a small structured count
matrix so the page remains self-contained.

The workflow is:

1. load or generate a feature-by-cell count matrix;
2. depth-subsample UMIs at several fractions;
3. call `fit_umi_scaling()` to fit spectra across the depth grid;
4. compare theoretical and empirical normalized overlap scores;
5. use `predict()` to extrapolate to higher depth.

## Load real Jurkat counts

```{r load-counts}
counts_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_counts_hvg.csv"
)

if (file.exists(counts_file)) {
  counts_full <- as.matrix(read.csv(counts_file, row.names = 1, check.names = FALSE))
  counts_source <- "Jurkat 10x HVG block"
} else {
  counts_full <- make_tutorial_counts(seed = 11)
  counts_source <- "synthetic structured tutorial block"
}

counts_source
dim(counts_full)
summary(colSums(counts_full))
```

To keep rendering fast, use 3,000 cells from the 5,000-cell tutorial block.

```{r choose-cells}
cells_use <- sample(colnames(counts_full), 3000)
counts_base <- counts_full[, cells_use, drop = FALSE]
summary(colSums(counts_base))
```

## Fit the UMI-depth scaling object

`fit_umi_scaling()` binomially thins counts to each UMI fraction, compares each
downsampled spectrum to the full-depth reference with `mi_theory()`, and fits a
saturating curve to normalized MI.

```{r umi-fit}
umi_fit <- fit_umi_scaling(
  counts_base,
  umi_grid = c(0.02, 0.05, 0.1, 0.2, 0.4, 0.7, 1.0),
  n_cells = ncol(counts_base),
  n_features = 300,
  transform = "log1p",
  min_cells = 10,
  r = 8,
  R = 3,
  p_sim = 100,
  seed = 1
)

umi_fit$data
```

## Empirical reference overlap

For a direct check, compare each UMI-thinned RNA embedding with the full-depth
RNA embedding using the same subspace-overlap MI formula. We normalize both
theoretical and empirical MI as `1 - exp(-2 * MI / r)`, which puts the scores on
the same bounded overlap scale.

```{r umi-empirical-mi}
umi_grid <- c(0.02, 0.05, 0.1, 0.2, 0.4, 0.7, 1.0)

normalized_overlap <- function(mi, r) {
  1 - exp(-2 * mi / pmax(r, 1))
}

umi_embedding <- function(x, r = 8, n_features = 300, min_cells = 10) {
  features <- select_hvgs(x, n_features = n_features, min_cells = min_cells)
  right_singular_vectors(log1p(x[features, , drop = FALSE]), r = r)
}

thin_counts <- function(x, fraction) {
  matrix(
    rbinom(length(x), size = as.integer(x), prob = fraction),
    nrow = nrow(x),
    dimnames = dimnames(x)
  )
}

z_full <- umi_embedding(counts_base, r = 8, n_features = 300, min_cells = 10)

umi_empirical <- do.call(rbind, lapply(umi_grid, function(frac) {
  set.seed(1 + 10000L + as.integer(round(frac * 10000)))
  counts_sub <- thin_counts(counts_base, frac)
  z_sub <- umi_embedding(counts_sub, r = 8, n_features = 300, min_cells = 10)
  emp <- subspace_overlap_mi(z_sub, z_full)
  data.frame(
    umi_fraction = frac,
    mean_umi_per_cell = mean(colSums(counts_sub)),
    empirical_mi = emp$mi,
    empirical_r_eff = emp$r_eff,
    empirical_overlap = normalized_overlap(emp$mi, emp$r_eff)
  )
}))

umi_compare <- merge(umi_fit$data, umi_empirical, by = "mean_umi_per_cell")
umi_compare$theory_overlap <- normalized_overlap(umi_compare$mean_mi_norm, 1)
umi_compare[, c("mean_umi_per_cell", "theory_overlap", "empirical_overlap", "mean_mi_norm", "empirical_mi", "empirical_r_eff")]
```

## Inspect and extrapolate

```{r inspect-umi-fit}
coef(umi_fit)
summary(umi_fit)
predict(umi_fit, data.frame(mean_umi_per_cell = c(2500, 3500, 5000)))
```

## Plot theory scaling law

```{r plot-umi-law}
plot(umi_fit, xlab = "Mean UMI per cell in HVG block", ylab = "Theoretical MI / retained rank")
```

## Plot empirical overlap

```{r plot-umi-empirical}
plot(
  umi_compare$mean_umi_per_cell,
  umi_compare$empirical_overlap,
  pch = 17,
  col = "#d95f02",
  type = "b",
  xlab = "Mean UMI per cell in HVG block",
  ylab = "Empirical normalized overlap"
)
```

## Adapt to real data

Use another real count matrix as `counts_full`, repeat the binomial thinning
across fractions and replicates, and store the response you want to model. The
spectral estimation pieces stay the same.
