---
title: "Cell number scaling law"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "figure/cell-number-"
)

local_lib <- normalizePath(file.path("..", "..", "..", ".rlib"), mustWork = FALSE)
if (dir.exists(local_lib)) .libPaths(c(local_lib, .libPaths()))
library(scScale)
set.seed(11)
```

## Goal

This example asks how a spectral signal estimate changes as the number of cells
increases. It starts from a real Jurkat 10x single-cell count block already
sampled in this repository: 400 highly variable genes by 5,000 cells, derived
from `data/Jurkat/sample_filtered_feature_bc_matrix`.

The workflow is:

1. load the real feature-by-cell count matrix;
2. sample different numbers of cells;
3. call `fit_cell_scaling()` to fit spectra across the grid;
4. inspect `coef()`, `summary()`, and `plot()`;
5. use `predict()` to extrapolate to larger cell counts.

## Load real Jurkat counts

```{r load-counts}
counts_file <- file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_counts_hvg.csv"
)

counts <- as.matrix(read.csv(counts_file, row.names = 1, check.names = FALSE))
dim(counts)
summary(colSums(counts))
```

The full source dataset contains many more cells; this tutorial uses a compact
HVG block so it renders quickly.

```{r source-summary}
sample_summary <- read.csv(file.path(
  "..", "..", "..",
  "outputs", "exploration", "jurkat_glmpca_gaussian_snr_by_n",
  "jurkat_glmpca_sample_summary.csv"
))
sample_summary[, c("dataset", "n_total_cells", "sampled_cells", "n_genes_used", "mean_umi_per_cell")]
```

## Fit the cell-number scaling object

`fit_cell_scaling()` computes the reference spectrum from the full matrix,
subsamples cells across `cell_grid`, computes theoretical MI with `mi_theory()`,
and fits a saturating curve to normalized MI.

```{r cell-fit}
cell_fit <- fit_cell_scaling(
  counts,
  cell_grid = c(250, 500, 1000, 2000, 3500, 5000),
  n_features = 300,
  transform = "log1p",
  min_cells = 10,
  r = 8,
  R = 3,
  p_sim = 100,
  seed = 1
)

cell_fit$data
```

## Inspect and extrapolate

The fitted object supports the usual R-style methods.

```{r inspect-cell-fit}
coef(cell_fit)
summary(cell_fit)
predict(cell_fit, data.frame(n_cells = c(7500, 10000, 20000)))
```

## Plot

```{r plot-cell-law}
plot(cell_fit, xlab = "Cells sampled", ylab = "Theoretical MI / retained rank")
```

## Adapt to real data

Replace `counts` with another real feature-by-cell matrix or Seurat object
accepted by `find_eigenvalues()`. Keep the same cell grid loop, then choose the
response that best matches the biological question.
